---
title: "Chapter 2 - Summarizing Data"
author: ""
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(openintro)
```

**Stats scores**. (2.33, p. 78) Below are the final exam scores of twenty introductory statistics students. 

57, 66, 69, 71, 72, 73, 74, 77, 78, 78, 79, 79, 81, 81, 82, 83, 83, 88, 89, 94

Create a box plot of the distribution of these scores. The five number summary provided below may be useful.

\begin{center}
\renewcommand\arraystretch{1.5}
\begin{tabular}{ccccc}
Min & Q1    & Q2 (Median)   & Q3    & Max \\
\hline
57  & 72.5  & 78.5          & 82.5  & 94 \\
\end{tabular}
\end{center}

```{r, echo = FALSE}
scores <- c(57, 66, 69, 71, 72, 73, 74, 77, 78, 78, 79, 79, 81, 81, 82, 83, 83, 88, 89, 94)



ggplot(as.data.frame(scores), aes(x="", y=scores)) + 
geom_boxplot() +
coord_cartesian(ylim = c(40, 120)) +
xlab("Box Plot of 20 grades") 



```





--------------------------------------------------------------------------------

\clearpage

**Mix-and-match**. (2.10, p. 57) Describe the distribution in the histograms below and match them to the box plots.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}

set.seed(7365)
sym = rnorm(1000, mean = 60, sd = 3)
uni = runif(1000, min = 0, max = 100)
rs = rgamma(1000, shape = 3, rate = 2)


# las label style = 1 = horizontal
# mfrow = tells the environment to position 6 plots in 1 row in the figure
# mgp = margin line for title, label and line ( default is 3,1,0 )
# mar = lines of margin on each asix, in the form of c(bottom, left, top, right)
# cex.lab    The magnification  for x and y labels
# cex.axis   The magnification for axis annotation

par(mar=c(4, 3.6, 0, 0), las = 1, mgp = c(2.7, 0.7, 0), 
    mfrow = c(1,6), 
    cex.lab = 1.5, cex.axis = 1.5)

# axes=FALSE means do not draw the X and Y tickers but note the next line, the axis() function will do it 
histPlot(sym, xlab = "(a)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(50,70,10))    # 50,60,70
histPlot(uni, xlab = "(b)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(0,100,50))    # 0,50,100
histPlot(rs, xlab = "(c)", ylab = "", col = COL[1], axes = FALSE)
axis(1, seq(0,6,2))       # 0,2,4,6

boxPlot(rs, xlab = "(1)", ylab = "", col = COL[1,3])
boxPlot(sym, xlab = "(2)", ylab = "", col = COL[1,3])
boxPlot(uni, xlab = "(3)", ylab = "", col = COL[1,3])







```

>*Histogram(a) and Box Plot(2) represent a* \underline{Normal Distribtuion} *produced by the rnorm() function.*
*The mean is 60 and the standard deviation is 3. So a normal distribution predicts 68.2\% of all observations will fall between 57 and 63.*

>*Histogram(b) and Box Plot(3) represent a* \underline{Uniform Distribution} *of 1000 numbers from 0 to 100.*
*Unlike the sample() function which returns discrete random numbers, runif() returns continous random numbers.*


>*Histogram(c) and Box Plot(1) reflect a* \underline{Gamma Distribution} *which is defined as a Continuous Probability Distribution that*
*models variables that are always positive and have skewed distributions. An example might be the wait time while standing on a line.*

>*The shape and scale parameter, often referred to as alpha and theta, effect the curvature of the curve, as opposed to shifting the curve*
*the way that standard deviation or mean would do.*


**Note:**
*All "r" functions produce a random set of numbers that theoretically fit the parameters of the given distribution.*
*However the seed function fixes the intitial "random" element so the above code will always return the same results when run again and again.*



--------------------------------------------------------------------------------

\clearpage

**Distributions and appropriate statistics, Part II**. (2.16, p. 59) For each of the following, state whether you expect the distribution to be symmetric, right skewed, or left skewed. Also specify whether the mean or median would best represent a typical observation in the data, and whether the variability of observations would be best represented using the standard deviation or IQR. Explain your reasoning.

(a) Housing prices in a country where 25% of the houses cost below $350,000, 50% of the houses cost below $450,000, 75% of the houses cost below $1,000,000 and there are a meaningful number of houses that cost more than $6,000,000.
(b) Housing prices in a country where 25% of the houses cost below $300,000, 50% of the houses cost below $600,000, 75% of the houses cost below $900,000 and very few houses that cost more than $1,200,000.
(c) Number of alcoholic drinks consumed by college students in a given week. Assume that most of these students don't drink since they are under 21 years old, and only a few drink excessively.
(d) Annual salaries of the employees at a Fortune 500 company where only a few high level
executives earn much higher salaries than the all other employees.



>*(a) Right skewed. The median is around 450,000 but the mean will be greater. I suppose the median is more reflective of the average home-owner.*

>*The mean isnt useful by itself because it cant distinguish between 10 homes worth 6MM or 1 home worth 60MM.*
>*The IQR is useful if you specifically know what Q1 and Q3 are. Youve already told me that Q1=350K and Q3=1MM and that tells me a lot.*
>*in terms of how I visualize the economy of that country.*
>*Standard deviation is only useful if you also know the mean. You also want to know if the curve is skewed since*
>*it is weighted equally by values below and above the mean.*

>*(b) Normalized. I already know the mean and the IQR. It sounds like very few outliers but the mean would be useful to know.*

>*If the mean greately exceeded the median, it would  tell me that there were some very expensive homes.*
         
>*(c) Right skewed. In fact it sounds like Q1, Q2 and beyond are 0. Given that, I would like to know the IQR to understand Q3.*

>*The mean would be useful to get a sense of how much the drinkers are drinking.*

>*(d) Right skewed. In this case, I dont know much except a few outliers exist on the right tail (upper tail).*

>*I would like to know the median and that is because I feel I would understand the salaries of most employees better*
          *The median will not be altered by the outliers whereas the outliers will weigh up the mean.*
          *Similarly, the IQR will not be effected by the few executives. If I know the median, and the IQR range, I will understand*
          *the salaries of most employees.*
          

**Note:**
*The statistic that is most helpful depends upon the context of what you already know and what you are trying to find out.*



--------------------------------------------------------------------------------

\clearpage

**Heart transplants.** (2.26, p. 76) The Stanford University Heart Transplant Study was conducted to determine whether an  experimental heart transplant program increased lifespan. Each patient entering the program was designated an official heart transplant candidate, meaning that he was gravely ill and would most likely benefit from a new heart. Some patients got a transplant and some did not. The variable *transplant* indicates which group the patients were in; patients in the treatment group got a transplant and those in the control group did not. Of the 34 patients in the control group, 30 died. Of the 69 people in the treatment group, 45 died. Another variable called *survived* was used to indicate whether or not the patient was alive at the end of the study.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=3}
library(openintro)
data("heart_transplant")



nrow(subset(heart_transplant,survived=="dead" & transplant=="control"))
# nrow(subset(heart_transplant,survived=="alive" & transplant=="control"))


# las label style = 1 = horizontal
# mgp = margin line for title, label and line ( default is 3,1,0 )
# mar = lines of margin on each asix, in the form of c(bottom, left, top, right)

# mosaic plot -------------------------------------------------------
par(mar = c(0, 0, 0, 0), las = 1, mgp = c(2.7, 0.9, 0))

# transplant is either control or treatment
# survived is either dead or alive
mosaicplot(transplant ~ survived, data = heart_transplant, 
           main = "", xlab = "", ylab = "", color = COL[1],
           cex.axis = 1.5)

# box plot ----------------------------------------------------------
par(mar = c(2, 4.8, 0, 0), las = 1, mgp = c(3.5, 0.7, 0), 
    cex.lab = 1.5, cex.axis = 1.5)
boxPlot(heart_transplant$survtime, fact = heart_transplant$transplant, 
        ylab = "Survival Time (days)", col = COL[1,2])
```

(a) Based on the mosaic plot, is survival independent of whether or not the 
patient got a transplant? Explain your reasoning.

*I would say tentatively there is a dependency. The survival rate increased from 65% to 88% with treatement.*
          *However, the sample size is not that great, and medical treatments are inheritently unpredictable.*
      

(b) What do the box plots below suggest about the efficacy (effectiveness) of the heart transplant treatment.

*The box plots reveal that the treatment also improves survival time in those patients who eventually died.*
          *Its not clear what it means when survived="alive" and yet survtime has a number.*


(c) What proportion of patients in the treatment group and what proportion of 
patients in the control group died?
```{r, comment=""}
treated_died<-nrow(subset(heart_transplant,survived=="dead" & transplant=="treatment"))
treated_total<-nrow(subset(heart_transplant, transplant=="treatment"))

untreated_died<-nrow(subset(heart_transplant,survived=="dead" & transplant=="control"))
untreated_total<-nrow(subset(heart_transplant, transplant=="control"))

sprintf("The percentage of treated patients who died is %.2f", treated_died/treated_total )
sprintf("The percentage of untreated patients who died is %.2f", untreated_died/untreated_total )

```



(d) One approach for investigating whether or not the treatment is effective 
is to use a randomization technique.

i. What are the claims being tested?
ii. The paragraph below describes the set up for such approach, if we were 
to do it without using statistical software. Fill in the blanks with a number 
or phrase, whichever is appropriate.

> We write *alive* on ___ 28 _______ cards representing patients who were 
alive at the end of the study, and *dead* on ___ 75 ______ cards 
representing patients who were not. Then, we shuffle these cards and split them 
into two groups: one group of size ____ 69 _____ representing treatment, and 
another group of size ____ 34 ______ representing control. We calculate the 
difference between the proportion of *dead* cards in the treatment and 
control groups (treatment - control) and record this value. We repeat this 100 
times to build a distribution centered at __ 0 ___. Lastly, we calculate 
the fraction of simulations where the simulated differences in proportions are 
___ 0.23 ______. If this fraction is low, we conclude that it is unlikely to 
have observed such an outcome by chance and that the null hypothesis should 
be rejected in favor of the alternative.

iii. What do the simulation results shown below suggest about the effectiveness 
of the transplant program?

```{r fig.height=3, fig.show="hold", message=FALSE, warning=FALSE, include=FALSE, out.width="50%", results='hide'}
# randomization -----------------------------------------------------

# creates summary of treatment/control and alive/dead 
# it notes the difference between treatement-success and non-treatment success is -0.2302
# then it returns a simulated difference, 100 trials
diffs <- DATA606::inference(heart_transplant$survived, heart_transplant$transplant, 
                  success = "dead", order = c("treatment","control"), 
                  est = "proportion", type = "ht", method = "simulation", 
                  nsim = 100, null = 0, 
                  alternative = "twosided", 
                  simdist = TRUE,
                  seed = 95632
                  )

# how do i see what inference does ?
# ?DATA606::inference

```

\begin{center}
```{r, results='hide', echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="75%", fig.height=3}
# plot randomization distribution -----------------------------------
par(mar = c(3.6, 2.2, 1, 1), las = 1, mgp = c(2.5, 0.7, 0), 
    cex.axis = 1.5, cex.lab = 1.5)

mean(diffs)

values  <- table(diffs)


plot(diffs, type = "n", xlim = c(-0.25, 0.25), 
     xlab = "simulated differences in proportions", 
     ylab = "", axes = FALSE, ylim = c(0, max(values)))
axis(1, at = seq(-0.25, 0.25, 0.05), 
     labels = c(-0.25, NA,-0.15, NA,-0.05, NA, 0.05, NA, 0.15, NA, 0.25))
for(i in 1:length(diffs)){
  x   <- diffs[i]
  rec <- sum(diffs == x)
  points(rep(x, rec), 1:rec, pch = 20, cex = 0.8, col = COL[1])
}
```
\end{center}



*The actuatl trial revealed a difference of .23, while the simulation did not reveal one difference that was as large as .23*

**Note:**
*Im not clear how this simulation incorporates a standard deviation that is appropriate to this medical treatment.*




