---
title: "Introduction to R and RStudio"
author: "Tom Buonora"
output: pdf_document
#output:
#  pdf_document: default
#  html_document:
#    includes:
#      in_header: header.html
#    css: ./lab.css
#    highlight: pygments
#    theme: cerulean
#    toc: true
#    toc_float: true
#editor_options: 
#  chunk_output_type: console
---

```{r global_options, include=FALSE}
# knitr::opts_chunk$set(eval = TRUE, results = FALSE, fig.show = "hide", message = FALSE, warning = FALSE)
library(tidyverse)
library(openintro)
```


## Dr. Arbuthnot's Baptism Records

To get started, let's take a peek at the data.

```{r load-abrbuthnot-data, eval=TRUE}
source('C:/Users/arono/Documents/R/win-library/4.0/DATA606/labs/Lab1/more/arbuthnot.r')
```

```{r view-data}
arbuthnot
```


```{r glimpse-data}
# apparently glimpse has been imported through 2 packages, tibble and dplyr
glimpse(arbuthnot)
```



```{r glimpse-data-result}
glimpse(arbuthnot)
```


## Some Exploration



```{r view-boys}
head(arbuthnot$boys)
```


1. What command would you use to extract just the counts of girls baptized? Try  it!
    
```{r extract-girls-only}
head(arbuthnot$girls)

```
    


### Data visualization

R has some powerful functions for making graphics. We can create a simple plot of the number of girls baptized per year with the command

```{r plot-girls-vs-year}
ggplot(data = arbuthnot, aes(x = year, y = girls)) +  geom_point()

```

Replace `geom_point()` with `geom_line()`.

```{r plot-girls-vs-year-line}
ggplot(data = arbuthnot, aes(x = year, y = girls)) + 
  geom_line()
```


1. Is there an apparent trend in the number of girls baptized over the years? 
   How would you describe it? (To ensure that your lab report is comprehensive,
   be sure to include the code needed to make the plot as well as your written
   interpretation.)

\textcolor{blue}{The overall trend shows an increase in the number of girls getting baptised each year. 
But there were some dramatic drops as well around 1640 and 1720, and some dramatic increaeses around 1660 and 1722.}

\textcolor{blue}{The below chunk calculates the delta of girls baptised each year and then prints out the top 5 years of increase and decrease.}

```{r}

arbuthnot <- arbuthnot %>%
  mutate(delta_girls = 0)

for (i in 2:nrow(arbuthnot))
{
  if (i==2)
  {
    prev_girls=arbuthnot[1,"girls"]
  }
  
  
  
  current_girls = arbuthnot[i,"girls"]
  delta_girls = current_girls-prev_girls
  arbuthnot[i,"delta_girls"]<-delta_girls
  
  prev_girls=current_girls
}

arbuthnot[which.max(arbuthnot$delta_girls),]
arbuthnot[which.min(arbuthnot$delta_girls),]

arbuthnot_sorted<-arbuthnot %>% arrange(desc(delta_girls))
print ("The greatest increaseses : ")
head(arbuthnot_sorted)
print("The greatest decreases")
tail(arbuthnot_sorted)
  
```

\textcolor{blue}{I was close. It was actually 1704 and 1705 which had the most dramatic deltas.}



Display rows where girls are greater than boys

```{r}
nrow(subset(arbuthnot, girls > boys))
```
Now display rows where boys are greater than girls
```{r}
nrow(subset(arbuthnot, boys > girls))

```

\textcolor{blue}{I would like to verify if we should assume that this ratio in baptism records actually infers the same ratio in actual births.}

### R as a big calculator
 
If we add the vector for baptisms for boys to that of girls, R will compute all sums simultaneously.

```{r calc-total-bapt-vars}
arbuthnot$boys + arbuthnot$girls
```


### Adding a new variable to the data frame

We'll be using this new vector to generate some plots, so we'll want to save it as a permanent column in our data frame.

```{r calc-total-bapt-vars-save}
arbuthnot <- arbuthnot %>%
  mutate(total = boys + girls)
```


```{r plot-total-vs-year}
ggplot(data = arbuthnot, aes(x = year, y = total)) + 
  geom_line()
```

Create a boy to girl ratio

```{r calc-prop-boys-to-girls-vars}
arbuthnot <- arbuthnot %>%
  mutate(boy_to_girl_ratio = boys / girls)
```


Create a boy to total ratio

```{r calc-prop-boys-vars}
arbuthnot <- arbuthnot %>%
  mutate(boy_ratio = boys / total)
```

*Note*
**mutate()** *adds new variables and preserves existing ones.*
**transmute()** *adds new variables and drops existing ones.*


3. Now, generate a plot of the proportion of boys born over time. What do you see? 


```{r plot-boys-total-ratio}
ggplot(data = arbuthnot, aes(x = year, y = boy_ratio)) + 
  geom_line()
```

\textcolor{blue}{My comment is that the graph seems choppy, but only because the scale is so narrow. I would characterize the male to female birth ratio as consistent within the 51 to 53 percent range.}



```{r boys-more-than-girls}
arbuthnot <- arbuthnot %>%
  mutate(more_boys = boys > girls)
```



## More Practice


```{r summarize min and max}
arbuthnot %>%
  summarize(min = min(boys), max = max(boys))
```

1.  What years are included in this data set? What are the dimensions of the 
    data frame? What are the variable (column) names?
    
    
\textcolor{blue}{Since there are 82 rows, I will just print the minimum/maximum of year}

```{r dim-present}

# arbuthnot["year"]


present <- present %>%  mutate(total = boys + girls)
present <- present %>%  mutate(boy_to_girl_ratio = boys / girls)
present <- present %>% mutate(more_boys = boys > girls)


sprintf("This study spanned the years from %s to  %s ", min(present["year"]),max(present["year"])) 

d_df <- dim(present)
sprintf("The dimensions of the table are %d rows and %d columns ", d_df[1],d_df[2])

cnames<-as.data.frame(names(present))
knitr::kable(cnames, caption='Column Names',col.names = "")

```
    
    
    
    

1.  How do these counts compare to Arbuthnot's? Are they of a similar magnitude?


\textcolor{blue}{The counts are signficantly higher. The actual numbers will be shown in a little bit when I calculate the mean totals}



```{r count-compare}


present_range<- max(present["year"]) - min(present["year"])

arbuthnot_range<- max(arbuthnot["year"]) - min(arbuthnot["year"])

sprintf("The arbuthnot years were %s to  %s or %d total years", 
        min(arbuthnot["year"]),max(arbuthnot["year"]), arbuthnot_range) 

sprintf("Meanwhile the present study covered %d years",present_range)

present_avg_total<-format(mean(present$total),big.mark = ",")

arbuthnot_avg_total<-format(mean(arbuthnot$total),big.mark = ",")

sprintf("The average total count was %s for arbuthnot and  %s for present ", 
          arbuthnot_avg_total, present_avg_total) 



sprintf("The average ration was %.4f for arbuthnot and  %.4f for present ", 
          mean(arbuthnot$boy_to_girl_ratio),  mean(present$boy_to_girl_ratio)) 




```



1.  Make a plot that displays the proportion of boys born over time. What do you see? 
    Does Arbuthnot's observation about boys being born in greater proportion than girls 
    hold up in the U.S.? 
    
    

```{r plot-prop-boys-present}

ggplot(data = present, aes(x = year, y = boy_to_girl_ratio)) +  geom_line()



```
    
    
\textcolor{blue}{I see the same ratio which, to me, is strong evidence that the ratio is natural to our species.}

    
1.  In what year did we see the most total number of births in the U.S.? 


```{r find-max-total}
# present_totals<-present %>% arrange(desc(total))

present[which.max(present$total),]


```

    
* * *
