---
title: "Introduction to linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, results = TRUE, fig.show = "show", message = FALSE, warning = FALSE)
```

The Human Freedom Index is a report that attempts to summarize the idea of "freedom"
through a bunch of different variables for many countries around the globe.
It serves as a rough objective measure for the relationships between the different
types of freedom - whether it's political, religious, economical or personal freedom -
and other social and economic circumstances. The Human Freedom Index is an annually
co-published report by the Cato Institute, the Fraser Institute, and the Liberales
Institut at the Friedrich Naumann Foundation for Freedom.

In this lab, you'll be analyzing data from Human Freedom Index reports from 2008-2016.
Your aim will be to summarize a few of the relationships within the data both 
graphically and numerically in order to find which variables can help tell a story
about freedom.



```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```


### The data

The data we're working with is in the openintro package and it's called `hfi`, 
short for Human Freedom Index.

1.  What are the dimensions of the dataset?
```{r}

sprintf("The hfi table has %d columns and %d rows", ncol(hfi), nrow(hfi))


```



1.  What type of plot would you use to display the relationship between the personal
    freedom score, `pf_score`, and one of the other numerical variables? Plot 
    this relationship using the variable `pf_expression_control` as the predictor (x).
    Does the relationship look linear? If you knew a country's `pf_expression_control`,
    or its score out of 10, with 0 being the most, of political pressures and 
    controls on media content, would you be comfortable using a linear model to
    predict the personal freedom score (y) ?
    
\textcolor{blue}{It looks somewhat linear although Im not sure if the wide range of the response variables is an issue for a linear model.}
    
```{r}

# note: expression control is somehat discrete and
#    one expression control value may map to many personal freedome scores
# View(hfi[c("year", "region", "countries", "pf_score", "pf_expression_control")])

hfi_new<-hfi %>%
  drop_na(pf_score) %>%
  drop_na(pf_expression_control)


x<-hfi_new$pf_score
y<-hfi_new$pf_expression_control


par(mfrow=c(1,2))
plot(x, y, main = "Scatter Plot", 
     xlab ="Personal Freedom Score", 
     ylab="Political pressures and controls on media content" )
abline(lm(y~x), col="red") # regression line (y~x)
lines(lowess(x,y), col="blue")   # non-linear best fit


# qqplot plots one against the other
qqplot(hfi$pf_score, hfi$pf_expression_control, xlab="expression control", ylab="personal freedom")


```
    
    

If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))


```

Here, we set the `use` argument to "complete.obs" since there are some observations
of NA.




## Sum of squared residuals

<div id="boxedtext">
In this section, you will use an interactive function to investigate what we mean by "sum 
of squared residuals". You will need to run this function in your console, not in your 
markdown document. Running the function also requires that the `hfi` dataset is loaded 
in your environment.
</div>

Think back to the way that we described the distribution of a single variable. 
Recall that we discussed characteristics such as center, spread, and shape. It's
also useful to be able to describe the relationship of two numerical variables, 
such as `pf_expression_control` and `pf_score` above.



1.  Looking at your plot from the previous exercise, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.

\textcolor{blue}{While I observe that they generally increase or decrease together, there are quite a few exceptions.}

Just as you've used the mean and standard deviation to summarize a single variable, 
you can summarize the relationship between these two variables by finding the 
line that best follows their association. Use the following interactive 
function to select the line that you think does the best job of going through 
the cloud of points.

```{r plotss-expression-score, eval=FALSE}
 # x = hfi_new$pf_score

plot_ss(x = x, y = y, showSquares = FALSE, leastSquares = FALSE) 

# View(plot_ss)

```

After running this command, you'll be prompted to click two points on the plot 
to define a line. Once you've done that, the line you specified will be shown in
black and the residuals in blue. Note that there are 30 residuals, one for each 
of the 30 observations. Recall that the residuals are the difference between the
observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes
the sum of squared residuals. To visualize the squared residuals, you can rerun 
the plot command and add the argument `showSquares = TRUE`.

```{r plotss-expression-score-squares, eval=FALSE}

plot_ss(x = x, y = y, showSquares = TRUE)


```

Note that the output from the `plot_ss` function provides you with the slope and
intercept of your line as well as the sum of squares.

1.  Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?

\textcolor{blue}{After running it 3 times, I have the following linear equations.}
\textcolor{blue}{The last one is a bit high. The intercept seems to have more effect than I expected.}

$$y=1.33x - 3.283  \ \ \ \ \ \  (SSE \ or \ RSS = 4054.05)$$   
$$y=1.271x - 2.96 \ \ \ \ \ \  (SSE \ or \ RSS = 3697.986)$$   

$$y=1.279x - 2.598 \ \ \ \ \ \  (SSE \ or \ RSS = 5020.977)$$ 

$$y=1.335x - 2.462 \ \ \ \ \ \  (SSE \ or \ RSS = 7448.953)$$ 



## The linear model

It is rather cumbersome to try to get the correct least squares line, i.e. the 
line that minimizes the sum of squared residuals, through trial and error. 
Instead, you can use the `lm` function in R to fit the linear model (a.k.a. 
regression line).

```{r m1}

# NOTE: y ~ x 
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)

plot(hfi_new$pf_expression_control,hfi_new$pf_score)


```

.

```{r summary-m1}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe 
the model is shown at the top. After the formula you find the five-number 
summary of the residuals. 





The "Coefficients" table shown next is key; its first 
column displays the linear model's y-intercept and the coefficient of `pf_expression_control`.
With this table, we can write down the least squares regression line for the 
linear model:

\[
  \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control
\]


```{r}
s1<-summary(m1)

# View(s1$residuals)
# View(resid(s1))

# the second row is the responnse variable pf_expression control
coef(s1)
s1$coefficients[2,]


s1_b<-s1$coefficients[1,1]
s1_m<-s1$coefficients[2,1]
RSS<-deviance(m1)    # RSS

sprintf("y = %.3f * x + %.3f        RSS=%.3f",s1_b, s1_m, RSS )



```


One last piece of information we will discuss from the summary output is the 
Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the 
proportion of variability in the response variable that is explained by the 
explanatory variable. For this model, 63.42% of the variability in pf_expression_control (y) is 
explained by pf_score(x)

```{r}
R<-cor(hfi_new$pf_score, hfi_new$pf_expression_control)
sprintf('R is %.2f. Since its positive, above .5, it tells me as x increases, y increases',R)
sprintf('R Squared is %.2f. R Squared is reffered to as the variablity of y explained by x. ',R^2)

```

\textcolor{blue}{R squared obviously is the same metric as correlation R, but its the absolute magnitude of the correlation.}


1.  Fit a new model that uses `pf_expression_control` to predict `pf_score`, or
    the total human freedom score. Using the estimates from the R output, write 
    the equation of the regression line. What does the slope tell us in the 
    context of the relationship between human freedom and the amount of political
    pressure on media content?
    
    
```{r flip x and y}


m2 <- lm(pf_expression_control ~ pf_score, data = hfi)
s2<-summary(m2)

s2_b<-s1$coefficients[1,1]
s2_m<-s1$coefficients[2,1]
RSS2<-deviance(m2)    # RSS

sprintf("y = %.3f * x + %.3f        RSS=%.3f",s2_b, s2_m, RSS2 )

# plot(m2)

```
    

## Prediction and prediction errors

Let's create a scatterplot with the least squares line for `m1` laid on top.

```{r reg-with-line}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

Here, we are literally adding a layer on top of our plot. `geom_smooth` creates
the line by fitting a linear model. It can also show us the standard error `se`
associated with our line, but we'll suppress that for now.

This line can be used to predict $y$ at any value of $x$. When predictions are 
made for values of $x$ that are beyond the range of the observed data, it is 
referred to as *extrapolation* and is not usually recommended. However, 
predictions made within the range of the data are more reliable. They're also 
used to compute the residuals.

1.  If someone saw the least squares regression line and not the actual data, how
    would they predict a country's personal freedom school for one with a 6.7 rating
    for `pf_expression_control`? Is this an overestimate or an underestimate, and
    by how much? In other words, what is the residual for this prediction?
    
\textcolor{blue}{There are 2 ways to do this}

```{r}

new_x<-6.7          # where x=pf_expression_control
new_y<-(new_x*s2_m) + s2_b

sprintf("Using are known slope and intercept, and given x=%.2f then y= %.2f", new_x, new_y)

# there is no actual 6.7 in the data but there are many 6.5 and 6.75
#  subset(hfi_new, pf_expression_control<6.77 & pf_expression_control>6.4)[c("pf_expression_control","pf_score")]

six.seven <- data.frame( pf_expression_control = 6.7 )
predicted.fit<-predict(m1, newdata = six.seven, interval = "prediction")

sprintf("The predict function used our linear model and predicted pf_control= %.2f", predicted.fit[1,1])


```
    
    
    
    

## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1) 
linearity, (2) nearly normal residuals, and (3) constant variability.

**Linearity**: You already checked if the relationship between `pf_score` and 
`pf_expression_control' is linear using a scatterplot. We should also verify this
condition with a plot of the residuals vs. fitted (predicted) values.

```{r residuals}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")


```

Notice here that `m1` can also serve as a data set because stored within it are 
the fitted values ($\hat{y}$) and the residuals. Also note that we're getting 
fancy with the code here. After creating the scatterplot on the first layer 
(first line of code), we overlay a horizontal dashed line at $y = 0$ (to help us
check whether residuals are distributed around 0), and we also reanme the axis 
labels to be more informative.

1.  Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between the two variables?  

</br>

\textcolor{blue}{The data is somewhat linear.}

\textcolor{blue}{x and y tend to increase together.}

\textcolor{blue}{There are some outliers.}

\textcolor{blue}{My concern is for a given value of pf_expression_control the pf_score can vary quite a bit.}



```{r}


lm_xy<-data.frame(y = m1$residuals + m1$fitted.values, residuals = m1$residuals, fitted = m1$fitted.values )
hfi_xy<-data.frame(y = hfi_new$pf_score, x = hfi_new$pf_expression_control )
merged_xy<-merge(lm_xy, hfi_xy, by="y")

# View(merged_xy)


show_everyone<-merged_xy %>%
  filter(x>4 & x<5)  %>%
  group_by(x) %>%
  summarize(min_y=min(y), max_y=max(y), mean_y= mean(y)) 


knitr::kable(show_everyone, caption='Range of pf_score for every pf_expression_control ')

# View(cbind(m1$residuals, m1$fitted.values, m1$residuals + m1$fitted.values))
# View(cbind(hfi_new$pf_score, hfi_new$pf_expression_control))


```

**Nearly normal residuals**: To check this condition, we can look at a histogram

```{r hist-res}
# .resid seems to be the same as m1$residuals
ggplot(data = m1, aes(x = m1$residuals)) +
  geom_histogram(binwidth = .03) +
  xlab("Residuals")
```

or a normal probability plot of the residuals.

```{r qq-res}


ggplot(data = m1, aes(sample = .resid)) +
  stat_qq()

# shapiro.test(m1$residuals)
# qqnorm(m1$residuals)

```

Note that the syntax for making a normal probability plot is a bit different 
than what you're used to seeing: we set `sample` equal to the residuals instead 
of `x`, and we set a statistical method `qq`, which stands for "quantile-quantile",
another name commonly used for normal probability plots.

1.  Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?

\textcolor{blue}{Its somewhat normal.}


</br>

**Constant variability**:

1.  Based on the residuals vs. fitted plot, does the constant variability condition 
    appear to be met?

\textcolor{blue}{It seems somewhat constant yes.}

* * *

## More Practice
<div class="oyo">

-   Choose another freedom variable and a variable you think would strongly
    correlate with it.. Produce a scatterplot of the two variables and fit a 
    linear model. At a glance, does there seem to be a linear relationship?

\textcolor{blue}{Lets look at Security and Safety vs Freedome of Religion}



```{r}


# pf_ss =Security and Safety and pf_religion = Religious freedom

# View(cbind(hfi$pf_association_assembly, hfi$pf_religion))
 
 hfi_new<-hfi %>%
  drop_na(pf_ss) %>%
  drop_na(pf_religion)
 
 x<-hfi_new$pf_religion
 y<-hfi_new$pf_ss
 
 
 
 # y = pf_association_assembly 
 # m3 <- lm(hfi_new$pf_association_assembly ~ hfi_new$pf_religion, data = hfi_new)

 
 R<-cor(x,y)
sprintf('R Squared is %.2f. This is the variablity of y explained by x',R^2) 


# plot_ss(x = hfi_new$pf_religion, y =hfi_new$pf_association_assembly, showSquares = FALSE, leastSquares = FALSE) 

 plot(x, y, main = "Scatter Plot", 
     xlab ="Freedom of Religion", 
     ylab="Security and Safety" )
abline(lm(y~x), col="red") # regression line (y~x)
lines(lowess(x,y), col="blue") 



```


\textcolor{blue}{Not a strong relationship at all.}

-   How does this relationship compare to the relationship between 
    `pf_expression_control` and `pf_score`? Use the $R^2$ values from the two 
    model summaries to compare. Does your independent variable seem to predict
    your dependent one better? Why or why not?
    
\textcolor{blue}{Just by looking, it seems like there is no correlation at all. The R squared value before was .64 and this one is only .1}
\textcolor{blue}{Im not sure I expected a correlation although I think of some countries with little freedom have a lot of civil strife.}
    


-   What's one freedom relationship you were most surprised about and why? Display
    the model diagnostics for the regression model analyzing this relationship.

\textcolor{blue}{I find it difficult to assume that any of the metrics are clearly explanatory or response.}
\textcolor{blue}{It seems to me any 2 attributes may correlate without implying causality.}
\textcolor{blue}{Anyway originally I thought the pf score correlation was weak, but as I continued to plot I changed my mind.}
\textcolor{blue}{The box plot was a pretty useful visualation to establish the xy relationship.}
\textcolor{blue}{So was qqline.}

```{r}

par(mfrow=c(1,2))

boxplot(hfi$pf_score ~ hfi$pf_expression_control, data=hfi, main = "Box Plot",
         xlab="Political pressures and controls on media content", 
         ylab="Personal Freedom Score")


# qqnorm plots one vector agains its theoretical normalized quantiles
qqnorm(hfi$pf_score)
qqline(hfi$pf_score, col="red")    
qqline(hfi$pf_expression_control, col="blue")  


```
    
    
* * *
