---
title: "Multiple linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE,  fig.show = "show", message = FALSE, warning = FALSE)
```

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" by Hamermesh and Parker 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. 

Here, you will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.



```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
library("gridExtra")

```

You will be using the `ggpairs` function from this package later in the lab.

### Creating a reproducible lab report


### The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. The result is a data frame 
where each row contains a different course and columns represent variables about 
the courses and professors. It's called `evals`.

```{r, eval=FALSE}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and some 
numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-evals, eval=FALSE}
?evals
```

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.


\textcolor{blue}{The study is observational.}

\textcolor{blue}{It is a difficult question to answer since each evaluator will have a different perception of the teachers appearance.}

\textcolor{blue}{I would rephrase it as "Are student evaluations unduly biased by the physical appearance of the teacher?"}


2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?

```{r}
hist(evals$score)
```

\textcolor{blue}{Its left skewed. Most of the values are on the high end.}

\textcolor{blue}{I would think most students appreciate their teachers and give them good grades.}


3.  Excluding `score`, select two other variables and describe their relationship 
    with each other using an appropriate visualization.
    
\textcolor{blue}{Im curious how the beauty scores were distributed by age and gender.}
    
```{r}

ggplot(data = evals, aes(x = age, y = bty_avg, fill = gender)) +
  geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75)  +
  ylim(0,6) +
  labs(x = "\n Age", y = "Score\n", title = "\n Scores by Age and Gender \n") +
  theme(plot.title = element_text(hjust = 0.5),      # center the title
        axis.title.x = element_text(face="bold", colour="red", size = 12),
        axis.title.y = element_text(face="bold", colour="red", size = 12),
        legend.title = element_text(face="bold", colour="blue", size = 10))



```

\textcolor{blue}{Age didnt matter that much. This helps me to understand what is meant by beauty.}

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg}

ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()

```

Before you draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

4.  Replot the scatterplot, but this time use `geom_jitter` as your layer. What 
    was misleading about the initial scatterplot?

```{r scatter-score-bty_avg-jitter}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```

\textcolor{blue}{geom jitter enables 3 observations that fall on the same dot to appear as 3 dots close together.}

\textcolor{blue}{It sacrifices precision for the sake of a better overall view.}


5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating. Write out the equation for the linear 
    model and interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
Add the line of the bet fit model to your plot using the following:
    
```{r scatter-score-bty_avg-line-se}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")
```


```{r}
m_bty<-lm(score~bty_avg, data=evals)

sum_m_bty<-summary(m_bty)

sprintf("The r squared is %.3f which is pretty low so its not a great linear relationship.",
        sum_m_bty$r.squared)
sprintf("The residual standard error is %.3f which is pretty high, again showing for any given bty_avg there is a wide range of scores."
        ,sum_m_bty$sigma)

```




The blue line is the model. The shaded gray area around the line tells you about the 
variability you might expect in your predictions. To turn that off, use `se = FALSE`.

```{r scatter-score-bty_avg-line}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)
```

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).

\textcolor{blue}{The plot on the left does a better job showing that about .3 over the mean is very common.}
\textcolor{blue}{The plot on the right highlights the dispersion of the below average beauty averages}

```{r}




evals$residual <- resid(m_bty)



plot1<-ggplot(data = evals, aes(x = residual)) +
  geom_histogram(binwidth = .03) +
   ggtitle("Bty Avg Residual Freq Histogram") +
  xlab("Residuals") +
  theme(plot.title = element_text(hjust = 0.5))


plot2<-ggplot(evals, aes(x = bty_avg, y = residual)) +
  geom_hline(yintercept = 0) +
  geom_point() +
  ggtitle("Resid Plot Zero Intercept") +
  xlab("Beauty Average Score") +
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(plot1, plot2, ncol = 2)


```
    
    
    

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))


```

As expected, the relationship is quite strong---after all, the average score is 
calculated using the individual scores. You can actually look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after you've accounted for the professor's gender, you can add the gender term 
into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
sum_m_bty_gen<-summary(m_bty_gen)
sum_m_bty_gen
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.


\textcolor{blue}{The conditions for least square regression are :}

    1)	Linearity
    2)	Normality of the Residuals
    3)	Constant Variability
    4)	Independence
    
    
\textcolor{blue}{Here is a scatterplot of residual variability ( each residual squared ).}

\textcolor{blue}{The most significant outlier was professer 30.}

\textcolor{blue}{That professor had a bueaty of 5.1 yet scored only 2.3}

```{r}


evals$residual_gen <- resid(m_bty_gen)

evals$residual_gen_var <- (evals$residual_gen)^2

plot(x=evals$residual_gen_var, main="Residual Scatterplot by Bty/Gender", 
     xlab="Observations ", ylab="Residual Squared", pch=19, frame=T)


```


\textcolor{blue}{Condition 2 looks ok as we saw earlier, its a bit left skewed. 3 and 4 look very good.}

\textcolor{blue}{The relationship between beauty and assessment score is weak.}

\textcolor{blue}{It seems to me using linearity as a condition for least squared regression is somewhat circular logic since the metrics
are in part designed to assess linearity.}

\textcolor{blue}{The R squared improved by adding gender.}

\textcolor{blue}{In both models the RSE was high, so neither relationship is very linear.}



8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
\textcolor{blue}{Reviewing the estimates, it seems adding gender adjusts the slop up a bit.}
    
```{r}

# lets put them in a table

est.df<-data.frame(predictor="Intercept",
                   bty=sum_m_bty$coefficients[1][1],
                   btygender=sum_m_bty_gen$coefficients[1][1]
                   )

est.df<-rbind(est.df, data.frame(predictor=labels(sum_m_bty_gen$terms)[1],
                                 bty=sum_m_bty$coefficients[2][1],
                                 btygender=sum_m_bty_gen$coefficients[2][1]
                                 ))

est.df<-rbind(est.df, data.frame(predictor=labels(sum_m_bty_gen$terms)[2],
                                 bty=NA,
                                 btygender=sum_m_bty_gen$coefficients[3][1]
                                 ))



knitr::kable(est.df, caption='Estimate Comparison')


```
    

\textcolor{blue}{Reviewing the output metrics, it seems adding gender improves our correlation.}

\textcolor{blue}{Reviewing the output metrics, r squared is higher, that infers a better linearity, better correlation.}

\textcolor{blue}{The Fstatistic, however, decreased which infers the overall residual increased compared to the total sum of squares.}



```{r}
out.df<-data.frame( metric="r.squared",
                   bty=sum_m_bty$r.squared, 
                   btygender=sum_m_bty_gen$r.square
                   )

out.df<-rbind(out.df, data.frame(metric="adj.r.squared",
                                  bty = sum_m_bty$adj.r.squared, 
                                 btygender = sum_m_bty_gen$adj.r.squared 
                                 ))

out.df<-rbind(out.df, data.frame(metric="rse",
                                  bty = sum_m_bty$sigma, 
                                 btygender = sum_m_bty_gen$sigma
                                 ))


out.df<-rbind(out.df, data.frame(metric="fstat",
                                  bty = sum_m_bty$fstatistic[1], 
                                 btygender = sum_m_bty_gen$fstatistic[1]
                                 ))

knitr::kable(out.df, caption='Model Output Comparison')

```
    
Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `male` and `female` to being an 
indicator variable called `gendermale` that takes a value of $0$ for 
female professors and a value of $1$ for male professors. (Such variables 
are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied by zero,
leaving the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

<!-- We can plot this line and the line corresponding to those with color pictures
with the following  -->
<!-- custom function. -->

<!-- ```{r twoLines} -->
<!-- ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) + -->
<!--   geom_smooth(method = "lm", formula = y ~ x, se = FALSE) -->
<!-- ``` -->

```{r}
ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) +
   geom_smooth(method = "lm", formula = y ~ x, se = FALSE)


```


9.  What is the equation of the line corresponding to those with color pictures? 
    (*Hint:* For those with color pictures, the parameter estimate is multiplied
    by 1.) For two professors who received the same beauty rating, which color 
    picture tends to have the higher course evaluation score?
    
\textcolor{blue}{Pictures that are black and white have a coefficient of 0.}

\textcolor{blue}{Whatever effect black and white photos have is built in to the other coefficients.}

```{r}
fit<-lm(score ~ bty_avg + pic_color, data = evals)
beta_0<-fit$coefficients[1][1]
beta_1<-fit$coefficients[2][1]
beta_2<-fit$coefficients[3][1]

sprintf("beta_0 = %.2f   beta_1 = %.2f    beta_2 = %.2f ", beta_0,beta_1,beta_2)

```
    





$$score=\begin{cases}
pic = color & \text{-0.16059pic + btyavg * 0.05548 +  4.06318}\\
pic = blackwhite &  \text{0 + btyavg * 0.05548 +  4.06318}
\end{cases}$$

```{r}


ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)))


```

The decision to call the indicator variable `gendermale` instead of 
`genderfemale` has no deeper meaning. R simply codes the category that 
comes first alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel()` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.
    
    
```{r}
m_bty_rank<-lm(score ~ bty_avg + rank, data = evals)

summary(m_bty_rank)

```
    

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
gender, ethnicity, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.

\textcolor{blue}{The gendermale pvalue is very low, I guess it really is important to the score}

\textcolor{blue}{Below I added the number of students coefficients to the previous table.}

```{r}
m_bty_std <- lm(score ~ bty_avg + cls_students, data = evals)
sum_m_std<-summary(m_bty_std)

# intercept
est_new_col.df<-data.frame(btystudents=sum_m_std$coefficients[1][1]  )

est_new_col.df<-rbind(est_new_col.df, data.frame(
                                 btystudents=sum_m_std$coefficients[2][1]
                                 ))

est_new_col.df<-rbind(est_new_col.df, data.frame(
                                 btystudents=NA
                                 ))

est_new.df<-cbind(est.df,est_new_col.df)



est_new_row.df<-data.frame(predictor="n_students",
                  bty=NA,
                  btygender=NA,
                   btystudents=sum_m_std$coefficients[3][1]
                   )

est_new.df<-rbind(est_new.df, est_new_row.df)

knitr::kable(est_new.df, caption='Estimate Comparison')
```









13. Interpret the coefficient associated with the ethnicity variable.

\textcolor{blue}{I repeated the exercise. In this case ethnicity (not minority) improved the score a little less than gender (male)}

```{r}
m_bty_std <- lm(score ~ bty_avg + ethnicity, data = evals)
sum_m_std<-summary(m_bty_std)

# intercept
est_new_col.df<-data.frame(btyethnicity=sum_m_std$coefficients[1][1]  )

est_new_col.df<-rbind(est_new_col.df, data.frame(
                                 btyethnicity=sum_m_std$coefficients[2][1]
                                 ))

est_new_col.df<-rbind(est_new_col.df, data.frame(
                                 btyethnicity=NA
                                 ))

est_new.df<-cbind(est.df,est_new_col.df)



est_new_row.df<-data.frame(predictor="ethnicity",
                  bty=NA,
                  btygender=NA,
                   btyethnicity=sum_m_std$coefficients[3][1]
                   )

est_new.df<-rbind(est_new.df, est_new_row.df)

knitr::kable(est_new.df, caption='Estimate Comparison')
```






14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?
    
\textcolor{blue}{The highest pvalue belonged the cls_profs when category level=multiple}
    
```{r m_almost_full}
m_almost_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_almost_full)
```
    
\textcolor{blue}{It didnt make a significant difference. The original coefficient was quite small.}    
    

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.

\textcolor{blue}{Ill prioritize adjusted r squared but also look at rse.}

    
```{r m_backwards}

m_backwards <- lm(score ~  rank + gender + ethnicity +  age   + language  + cls_perc_eval  
              + cls_students + bty_avg + cls_credits +  pic_color  + pic_outfit 
            , data = evals)
    
sprintf("New Model Adj R Squared = %.3f.   RSE = %.3f", 
        summary(m_backwards)$adj.r.squared, summary(m_backwards)$sigma)

sprintf("Full Less Gender Model Adj R Squared = %.3f.   RSE = %.3f", 
        summary(m_almost_full)$adj.r.squared, summary(m_almost_full)$sigma)

sprintf("Full Model Adj R Squared = %.3f.   RSE = %.3f", 
        summary(m_full)$adj.r.squared, summary(m_full)$sigma)

```



\textcolor{blue}{Seems like each time i remove something r-squared decreases and rse increases. I removed cls_profs and cls_level which helped a little}

\textcolor{blue}{I tried to do this programatically but it was taking too much time so i used notepad to format this...}

score = \\
4.0856255 + \\
ranktenure track      * -0.142 \\
ranktenured           * -0.089 \\
gendermale            * 0.203 \\
ethnicitynot minority * 0.142 
age                   * -0.008 
languagenon-english   * -0.209 
cls_perc_eval         * 0.005
cls_students          * 0.0003   
bty_avg               * 0.041
cls_creditsone credit * 0.473
pic_colorcolor        * -0.197 
pic_outfitnot formal  * -0.117   




    

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
    
    
\textcolor{blue}{The conditions for least square regression are :}

    1)	Linearity
    2)	Normality of the Residuals
    3)	Constant Variability
    4)	Independence
    
    
\textcolor{blue}{Here is a scatterplot of residual variability ( each residual squared ).}

\textcolor{blue}{My comments are the same as from question 7 which used a different equation.}


```{r}


evals$residual_b_gen <- resid(m_backwards)

evals$residual_gen_var <- (evals$residual_b_gen)^2

plot(x=evals$residual_gen_var, main="Final Model Residuals", xlab="Observations ", ylab="Residual Squared", pch=19, frame=T)





```
    

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?

\textcolor{blue}{Sure some class are relative low stress where no one gets upset.}

\textcolor{blue}{Other classes stressful on the students so maybe they are more critical in the evaluations.}


18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
\textcolor{blue}{It does appear the men received better grades ( by .2 )}

\textcolor{blue}{The model has any class with "some credits" receiving an extra .47 so I dont understand that.}

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
    
\textcolor{blue}{When it comes the bty_avg, I think the study reveals more about what the beauty assessors are thinking than anything else.}

\textcolor{blue}{The other attributes are more interesting. It is possible that age and gender influence students in general.}



* * *
