---
title: "Data 605: Week 1"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```



<br>
<font size="5" color="purple">Linear Transformations on a Generic Graphic</font>
<br>



<br>

This exercise builds a graphic using a matrix or table of coordinates that define smaller rectangles.

<br>

Then it applies a 2X2 transformation matrix to alter the graphic in several ways. It displays the animations in an X Window.

<br>

The X11 plots are drawn with fig.show='hide' so that this document does not display the many plots.

<br>


<br>

## Constants.

<br>

<br>
This constant controls the density of dots throught the entire plot.
<br>


```{r constants}

DEPTH_OF_DOTS<-50         # number of plot points within a unit

```


## Generic functions.

<br>
These 2 functions create arrays of x,y points that populate any box within a nXn grid, based on 4 parameters.
<br>



```{r pop_arrays}

# Every box can be built with 4 parameters, these 2 functions share parameters since they need to be conformable


pop_x_array<-function(beg,end,x_axis_repeats, y_axis_repeats) {
  
  return(c(rep(seq(beg,end,length.out = x_axis_repeats),y_axis_repeats)))
}


# the x_axis_repeats is imortant bcuz every x point needs a y point

pop_y_array<-function(beg,end,x_axis_repeats, y_axis_repeats) {
  
 return(sort(rep(seq(beg,end,length.out = y_axis_repeats),x_axis_repeats)))
}

```


<br>

This one applys a transformation matrix to an x,y array.

<br>


```{r plot_transformation}



plot_transformation<-function(s,x_array,y_array, main_title) {

  # get the abcd of the transformation matrix
a<-s[1,1]
b<-s[1,2]
c<-s[2,1]
d<-s[2,2]


# create empty vectors
x_new<-c()
y_new<-c()

for(i in 1:length(x_array)) {
  x_pt<-x_array[i]            # get the x,y points
  y_pt<-y_array[i]
  
   
  x_pt_new<-a * x_pt + b * y_pt
  y_pt_new<-c * x_pt + d * y_pt
  x_new<-append(x_new,x_pt_new)
  y_new<-append(y_new,y_pt_new)
  
}

  plot(x_new,y_new, xlim = c(-1,1), ylim = c(-1,1), main=main_title)
}


```


<br>

This one accepts a theta in degrees.

<br>

It converts it to <b><i>radians</i></b> to create the trig rotation matrix. 

<br> 

Then it invokes plot_transformation()

<br>


```{r}

plot_rotated<-function(degrees,x,y) {
  
  theta<-degrees*pi/180
  a<-cos(theta)
  b<-sin(theta)*-1
  c<-sin(theta)
  d<-cos(theta)

  s<-matrix(c(a, b, c, d), byrow = TRUE, ncol = 2)

  plot_transformation(s,x,y, "Rotation")

}

```




<br>

## Build the Graphic

<br>

We build a graphic by constructing seperate rectangles.


<br>
Set up all the x,y points that define the smaller rectangles. 
<br>


```{r build_main_image}


x_y_points<-c(0 , 0.35 , 0.8 , 0.65, 32,     # top of T
      0.15 , 0.2 , 0 , 0.65, 200,            # bottom of T
      0.65 , 0.9 , 0.8 , 0.64, 200,           # top of B
      0.65 , 0.9 , 0.48 , 0.32, 30,              # middle of B
      0.65 , 0.9 , 0.16 , 0, 30,
      0.65 , 0.7 , 0.64 , 0.48, 30,
      0.65 , 0.7 , 0.32 , 0.16, 30,
      0.85 , 0.9 , 0.64 , 0.48, 30,
      0.85 , 0.9 , 0.32 , 0.16, 30)

points_matrix <- matrix(x_y_points, byrow = TRUE, ncol = 5)

```



<br>
Loop through the table, and pass the values to the plotting function.
<br>

```{r display_main_image}


x_points<-c()
y_points<-c()

for(row in 1:nrow(points_matrix)) {
    x_beg<-points_matrix[row, 1]
    x_end<-points_matrix[row, 2]
    y_beg<-points_matrix[row, 3]
    y_end<-points_matrix[row, 4]
    y_axis_repeats<-points_matrix[row, 5]
    x_axis_repeats<-as.integer(abs(x_end-x_beg)*DEPTH_OF_DOTS)
  
    x_points_tmp<-pop_x_array(x_beg,x_end,x_axis_repeats, y_axis_repeats)
    y_points_tmp<-pop_y_array(y_beg,y_end,x_axis_repeats, y_axis_repeats)
    
   
    x_points<-c(x_points,x_points_tmp)
    y_points<-c(y_points,y_points_tmp)
}


plot(x_points,y_points,pch='.')

```





***

## Shear

***

<br>
Shearing is when axis is fixed and the other moves.
<br>
Either
<br>

$$\begin{bmatrix} 	1&0\\ \lambda&1\\		\end{bmatrix}$$

or

<br>
$$\begin{bmatrix} 	1&\lambda\\ 0&1\\		\end{bmatrix}$$
<br>

```{r shearing, fig.show='hide'}

x11(width = 10,  height = 5, title="Shearing", display = "", bg="bisque")

for (t in seq(0.5,-0.5,by=-.1)) {

s<- matrix(c(1, t, 0, 1), byrow = TRUE, ncol = 2)
Sys.sleep(.2)
plot_transformation(s,x_points,y_points, "Shear")

}

for (t in seq(0.5,-0.5,by=-.1)) {

s<- matrix(c(1, 0, t, 1), byrow = TRUE, ncol = 2)
Sys.sleep(.2)
plot_transformation(s,x_points,y_points, "Shear")

}


```



***

## Scaling

***


<br>
Scaling is when increase x or y but the direction stays the same.
<br>
Either
<br>


$$\begin{bmatrix} 	1&0\\ 0&\lambda\\		\end{bmatrix}$$

or

<br>
$$\begin{bmatrix} 	\lambda&0\\ 0&1\\		\end{bmatrix}$$


```{r fig.show='hide'}

x11(width = 10,  height = 5, title="Scaling", display = "", bg="light blue")


for (t in seq(1,1.5,by=.1)) {

s<- matrix(c(t, 0, 0, 1), byrow = TRUE, ncol = 2)
Sys.sleep(.2)
plot_transformation(s,x_points,y_points, "Scaling - Dilation")

}


for (t in seq(1.5, .5, by=-.1)) {

s<- matrix(c(t, 0, 0, 1), byrow = TRUE, ncol = 2)
Sys.sleep(.2)
plot_transformation(s,x_points,y_points, "Scaling - Contraction")

}

```





***

## Rotating

***

<br>

The trig paradigm is that we apply the following:

<br>

$$\begin{bmatrix} 	cos(\theta)&-sin(\theta)\\ sin(\theta)&cos(\theta)\\		\end{bmatrix}$$

<br>

This means as our a-d values approach 1 the b-c values approach zero and vica versa
<br>



<br>

The function requires radians. Our theta is in degrees to make it easier to understand.

<br>



```{r echo=TRUE, fig.show='hide'}
x11(width = 10,  height = 5, title="Rotation", display = "", bg="green")

for (theta in seq(0,360,by=10)) {

  Sys.sleep(.1)
  plot_rotated(theta,x_points,y_points)
  
}


```







***

## Projection

***

<br>
Projection infers that the original shape remains the same. Its just displaced somehow.
<br>
There are many articles that describe it in terms of camera lenses or other complex applications.
<br>
These simple transformation matrices produce a *reflection*.
<br>

<br>
This
<br>


$$\begin{bmatrix} 	1&0\\ 0&-1\\		\end{bmatrix}$$

and this

<br>
$$\begin{bmatrix} 	-1&0\\ 0&1\\		\end{bmatrix}$$

```{r projection, fig.show='hide'}

x11(width = 10,  height = 5, title="Projection/Reflection", display = "", bg="bisque")


s<- matrix(c(1, 0, 0, -1), byrow = TRUE, ncol = 2)
Sys.sleep(2)
plot_transformation(s,x_points,y_points, "Projection")


s<- matrix(c(-1, 0, 0, 1), byrow = TRUE, ncol = 2)
Sys.sleep(2)
plot_transformation(s,x_points,y_points, "Projection")




```






