---
title: "Data 605: Week 5"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


<style type="text/css">
                      .tab { margin-left: 60px; }
</style>




<br><br><br><br>

<font size="7" color="purple">Probabilities</font>
<br><br><br>


<br><br><br><br>


***

## 1) Bayes Formula

***

<b>Sensitivity</b>
<p class="tab">a test's ability to designate an individual with disease as positive.</p>

<b>Specificity</b> 
<p class="tab">a test's ability to designate an individual who does not have a disease as negative.</p>

<br><br>

A new test has a 96% sensitivity and a 98% specific.

<br>

Given a .001 prevalence rate, what is the probability that an individual who is reported as positive by the new test actually has the disease?

<br>


```{r}
p_prevalence<-.001
p_sensitivity<-.96
p_specificity<-.98

num<-p_sensitivity * p_prevalence
den<-(p_sensitivity * p_prevalence) + ((1-p_specificity) * (1-p_prevalence))


sprintf("The probability that a person with a positive test result actually has the disease is %.6f ",num/den)


  
  
```

<br>

<font size="4" color="purple">Given that for every 1000 people only 1 has the disease, the majority of positive results will be from people who are not actually infected.</font>

<br>

Health        |  Population  |  Tests Positive  
------------- | ------------ | ----------------- 
Infected      |     10,000   |       9,600       
Not Infected  |  9,990,000   |     199,800  
------------  | -----------  | ----------------- 
Total         | 10,000,000   |     209,400




<br><br><br>


***

## 2) Binomial

***



<br>

The probability of your organization receiving a Joint Commission inspection in any given month is .05.

<br>

What is the probability that, after 24 months, you received exactly 2 inspections?

<br>

```{r}

x=2             # number of successes we want to know the probability of
n=24
p=.05

round(dbinom(x, size=n, prob=p),4)
  
```

<br>
What is the probability that, after 24 months, you received 2 or more inspections?

<br>

What is the probability that your received fewer than 2 inspections? 

<br>



```{r}

x=seq(0,1)             # number of successes we want to know the probability of
n=24
p=.05

pr<-sum(dbinom(x, size=n, prob=p))

print(paste("The probability of 2 or more is ", round(1-pr,4)))


  
print(paste("The probability of 1 or less is ", round(pr,4)))

```

<br>

What is the standard deviation?

<br>

```{r}

n*p*(1-p)


```

<br>

What is the expected number of inspections you should have received? 

<br>

```{r}

n*p

```

<br><br><br>

***

## 3) Poisson

***


<br>
You are modeling the family practice clinic and notice that patients arrive at a rate of 10 per hour. 

<br>

What is the probability that exactly 3 arrive in one hour?
<br>


```{r}

round(dpois(x = 3, lambda = 10),4)
round(dbinom(3,10,1),4)

```


<br>

What is the probability that more than 10 arrive in one hour?

<br>

```{r}

round(1 - ppois(q = 10, lambda = 10),4)

```

<br>
How many would you expect to arrive in 8 hours?

<br>

<font size="4" color="purple">The expected vaue = the mean = units X lambda so 80</font>

$$\mu \ = \ t * \lambda $$

<br>

<font size="4" color="purple">        so 80</font>

<br><br>

What is the standard deviation of the appropriate probability distribution?

<br>

$$\sigma \ = \ \sqrt{\mu}$$

<font size="4" color="purple">so 8.944272</font>

<br>

If there are three family practice providers that can see 24 templated patients each day, what is the percent utilization and what are your recommendations?

<br>
<font size="4" color="purple">In a 10 hour day I would expect 100 patients.The capacity of the 3 providers is well below that and not sufficient to handle 100 patients.</font>
<br>

<br><br><br>

***

## 4) Hypergeometric

***

<br><br>



Your subordinate is accused of favoring nurses. 

15 are nurses and 15 are not. 

The accuser stated that there were 6 company-paid trips to Disney World. 

The supervisor sent 5 nurses and 1 non-nurse. 

What was the probability of that ? 

<br>


```{r}


N = 30     #  (Total Population)
k = 15     #  of success in population
m = N-k    #  of failure in the population

n = 6      #  of trials
x = 5      #  of successes you are quantifying the probability


round(dhyper(x, m, k, n),4)


```

<br>

How many nurses would we have expected your subordinate to send?

<br>


```{r}

n * (k / (k + m))


```




<br><br><br>


***

## 5) Geometric

***




<br>

$$P(X=x) \ = \  (1-P)^{n-1}P$$




<br>
$$P(X \le x) \ = \  1 - (1 - P^n)$$

$$\sigma \ = \ \sqrt{\frac{1-p}{p^2}}$$

<br>
Expected Value ( Mean ) =
<br>
$$E(X) \ = \ \frac{1}{p} $$

<br>

The probability of being seriously injured in a car crash in an unspecified location is about .1% per hour. 


A driver traverses the area for 1200 hours 

What is the probability that the driver will be seriously injured? 


```{r}

p=.001
n=1200

round(1 - (1-p)^n,4)


```



In the course of 15 months? 

```{r}

# its 100 hours per months so
n=1500

round(1 - (1-p)^n,4)

```


What is the expected number of hours that a driver will drive before being seriously injured? 

```{r}

1/p

```

<br>

Given that a driver has driven 1200 hours, what is the probability that he or she will be injured in the next 100 hours?

<br>
<font size="4" color="purple">The phrase "Given" implies that we are not looking for the probability of 1300 hours of no accidents, just 100 hours.</font>
<br>

```{r}

n=100

1 - (1-p)^n

```

<br><br><br>

***

## 6) Poisson

***

<br>


You are working in a hospital that is running off of a primary generator which fails about once in 1000 hours. 
    What is the probability that the generator will fail more than twice in 1000 hours? 



<br>

```{r}

p_fail<-1 - pbinom(2, 1000, .001) 
sprintf("The probability that the generator will fail more than twice is %.2f", p_fail)

```


<font size="4" color="purple">Note: pbinom and ppois are very similar, and return almost the same result (if the rate < 1) </font>

<br>
`pbinom(2, 1000, .001) `
`ppois(2, lambda=1)`

<br>
<font size="4" color="purple">You could also convert the time unit if you think it would improve the distribution.</font>

<br>
      What is the expected value?
<br>

<font size="4" color="purple"> .0001 </font>


<br><br><br>

***

## 7) Uniform

***

<br>

A surgical patient arrives. The waiting time is uniformly distributed from 0 to 30 minutes.

<br>

What is the probability that this patient will wait more than 10 minutes? 


```{r}

# all this is q/(max-min)
1-punif(q=10, min = 0, max = 30)

```




<br>
If the patient has already waited 10 minutes, what is the probability that he/she will wait at least another 5 minutes prior to being seen?
<br>
```{r}

#  "at least another 5" means 5 or more
1-punif(q=4, min = 0, max = 20)

```




What is the expected waiting time?

<br>

<font size="4" color="purple"> The midpoint, so 15 minutes. </font>


<br><br><br>



***

## 8) Exponential

***

<br>


Your hospital owns an old MRI, which has a manufacturer’s lifetime of about 10 years (expected value). 

Based on previous studies, we know that the failure of most MRIs obeys an exponential distribution. 

What is the expected failure time? 

$$\lambda  \ = \ \frac{1}{E(X)}$$

<br>

<font size="4" color="purple">The expected value is 10 and the rate is .1 </font>


<br>


What is the standard deviation?

<br>

<font size="4" color="purple">The standard deviation is equal to the mean or Expected Value so .1  </font>


<br>



What is the probability that your MRI will fail after 8 years? 
<br>

```{r}

round(pexp(8,.1),4)

```


Now assume that you have owned the machine for 8 years. 
Given that you already owned the machine 8 years, what is the probability that it will fail in the next two years?

```{r}

round(pexp(10,.1) - pexp(8,.1),4)

```



