---
title: "Quiz_Week1"
author: "Tom Buonora"
date: "January 2, 2021"
output: html_document
---

<style type="text/css">
body {
font-size: 18px;
color: blue;
}
pre {
  font-size: 12px;
  color: maroon;
}
</style>





<br><br>

1. Create a vector that contains 20 numbers. (You may choose whatever numbers you like, but make sure there are some duplicates.)

```{r Q1}

# Jet win total for the past 20 years
vector_num1 = c(2,7,4,5,5,10,4,8,6,8,11,9,9,4,10,4,10,6,9,10)
length(vector_num1)           # 20

```
<br><br>

2. Use R to convert the vector from question 1 into a character vector.


```{r Q2}


vector_char1 <- as.character(vector_num1)

class(vector_char1)            # character


```
<br><br>
3. Use R to convert the vector from question 1 into a vector of factors

```{r Q3}


vector_factor1<-factor(vector_char1)
class(vector_factor1)       # factor


```
<br><br>
4. Use R to show how many levels the vector in the previous question has.


```{r Q4}


# note: R puts 10 and 11 before 2 since we created it from characters
nlevels(vector_factor1)  # 9
levels(vector_factor1)   #  "10" "11" "2"  "4"  "5"  "6"  "7"  "8"  "9"


```
<br><br>
 5. Use R to create a vector that takes the vector from question 1 and performs on it the formula 3^2−4X+1.

```{r Q5}

(3*vector_num1^2) - (4*vector_num1)  + 1


```
<br><br>
6. Create a named list. 

```{r Q6}


jets_defense = list(dline=c("Williams","Fatukasi","Anderson"), linebackers =c("Hewitt","Jenkins","Langi","Williamson"), secondary= c("Poole","Austin","Maye","Desir") )

names(jets_defense)
jets_defense["dline"]



```
<br><br>
7. Create a data frame with four columns – one each character, factor (with three levels), numeric, and date. 

```{r Q7}

player = c("Norm Van Brocklin", "Matt Schaub", "Warren Moon", "Boomer Esiason")
era = c("Fifties","Current","Nineties","Nineties")
yards = c(554,527,527,522)
date = c(as.Date("1951-09-28"),as.Date("2012-11-18"), as.Date("1990-12-16"), as.Date("1996-11-10"))

 # note: my data frames converted to character by default so here i have to explicitly call factor()
qb_best_games <- data.frame(player,era = factor(era), yards, date)
```
<br><br>
8. Illustrate how to add a row with a value for the factor column that isn’t already in the list of levels.

```{r Q8}


# this requires to replace the era column with itself, but specify the levels with the new level added
lvls <- levels(qb_best_games$era)
lvls[length(lvls) + 1] <- 'Eighties'
qb_best_games$era <- factor(qb_best_games$era, levels = lvls)

qb_best_games<-rbind(qb_best_games,data.frame(player= "Phil Sims", era= factor("Eighties"), yards = 513, date= as.Date('1085-10-13')))

```
<br><br>
9. Show the code that would read in a CSV file called temperatures.csv

```{r Q9}

qb_games <- read.csv("C:\\Users\\arono\\Documents\\R\\SportsStats\\qb_best_games.csv")

head(qb_games,10)


```
<br><br>
 10. Use a loop to calculate the final balance, rounded to the nearest cent, in an account that earns 3.24% interest 
               compounded monthly after six years if the original balance is $1,500.
```{r Q10}


principal <- 1500
rate <- .0324
freq <- 12
years <- 6

for (rolls in 1:(freq*years))
{
  principal=principal*(1+(rate/freq))
}

sprintf("%.2f",principal)
  # format method perhaps not as functional as sprintf
format(principal, digits=6, nsmall=2)




```
<br><br>
11. Create a numeric vector of length 20 and then write code to calculate the sum of every third element of the vector you have created.

```{r Q11}
# Jet win total for the past 20 years
vector_num1 = c(2,7,4,5,5,10,4,8,6,8,11,9,9,4,10,4,10,6,9,10)

vector_num1%%3                          # all the values of vector_num1  mod 3
vector_num1[1:20%%3==0]                 # the values of every 3rd number, i.e.  4 10  6  9 10  6

sum(vector_num1[1:20%%3==0])            # 45 = 4+10+6+9+10+6

```
<br><br>

12. Use a for loop to calculate : $\sum_{i=1}^{10} 2^i$



```{r Q12}


x=2
sumx=0

for (i in 1:10)
{
  sumx=sumx+x^i            # series is {2,4,8,16...1024}
}
sumx                        # 2046



```
<br><br>
13. Use a while loop to accomplish the same task as in the previous exercise

```{r Q13}


x=2
sumx=0
i=1

while (i <= 10)
{
  sumx=sumx+x^i            # series is {2,4,8,16...1024}
  i=i+1
}

print(sumx)                        # 2046

```

<br><br>
14. Solve the problem from the previous two exercises without using a loop.

```{r Q14}
sumx<-sum(2^(1:10))
sumx
```

<br><br>

15. Bonus : Work on Equations

See [here](https://www.math.mcgill.ca/yyang/regression/RMarkdown/example.html) for more info



Inline mathematical material is set off by the use of single dollar-sign characters. This is better to integrate text and equations.



This (\$\sum_{i=1}^n X_i\$)  is inline:  $\sum_{i=1}^n X_i$

Display form uses 2 dollar signs. ( use backslash if you every want to display a dollar sign )

$$\sum_{i=1}^n X_i$$
<br><br>


Underline for subscript. Carot for superscript. If the expression is more than one letter,then use brackets
\$\$prin^{1+rate}\$\$

$$prin^{1+rate}$$

<br><br>

Doesnt matter if subscript or superscript comes first
\$\$X^2_{i,j}\$\$

$$X^2_{i,j}$$
<br><br>
\$\$X_{i,j}^2\$\$
$$X_{i,j}^2$$
<br><br>

Use \sqrt for square root
\$\$\sqrt{b^2 - 4ac}\$\$
$$\sqrt{b^2 - 4ac}$$
<br><br>

Displayed fractions are typeset using the \frac operator.

\$\$\frac{4z^3}{16}\$\$
$$\frac{4z^3}{16}$$
<br><br>
Calculate the sum of $2^i$ where i is 1:10

\$\$\sum_{i=1}^{10} 2^i\$\$
$$\sum_{i=1}^{10} 2^i$$

<br><br>

use \\left and \\right to generate full size parenthesis
\$\$\sum_{i=1}^{n}\left( \frac{X_i}{Y_i} \right)\$\$
$$\sum_{i=1}^{n}\left( \frac{X_i}{Y_i} \right)$$

Greek letters :

$$\alpha, \beta,  \gamma, \Gamma$$

$x \ge 15$
$x \le 15$

<br><br>
n choose k
$$n \choose k$$


<br><br><br><br><br><br>

<p style="color:gold;font-size:24px;font-weight:bold;text-align:center;text-decoration: underline;">More Notes</p>


<!-- <center><u>More Notes</u></center> -->

<br><br>

<!-- <p style="color:gold;font-size:24px;font-weight:bold;text-align:center;>More Notes</p> -->



<p style="color:black;">Notes About Latex</p>

<ol>
   <li>Matrixes are achieved via begin and end for vmatrix, matrix or array</li>
   <li>italics 1 asterix</li>
   <li>bold    2 asterixes</li>
   <li>subscript tilde </li>
   <li>superscript carot </li>
</ol>

<i>Note : Latex is but one method to format equations. MathML and MathJAX are 2 others</i>



<br><br><br><br>


<p style="color:black;">Notes About Font/Color Formatting</p>


<ol>
   <li>Markdown and Colab both can incorprate basic html but its painful, syntax errors will not be reported !
   <li> see https://www.w3schools.com/cssref/ for css reference
   <li>legacy tags like \<center\>\<u\>\<b\> work but styles and css files would be better
   <li> i only got the css classes working for "pre" and "body", not for r.code
   <li>also note while saving off your html is nice. Its the rmd file you want to save as that is the true source code.
</ol>
