---
title: "TomTest2"
author: "Tom Buonora"
professor: Jeffrey Nieman
date: "December 17, 2020"
output: html_document
rpubs: https://rpubs.com/BigApple/705610
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to R Markdown

This is my first R Markdown document for a R workshop prep for the CUNY MSDS program. 

I found a good cheat sheet to assist me [here](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)

Here are the class assignments...



```{r factorial, echo=TRUE}

#  do what factorial(12) does

fact<- 1

for ( x in 1:12)
{
  fact<-fact*x
}

print (fact)
```




```{r sequence}
# print every 5 numbers
every5<-seq(from=20,to=50,by=5) 
print(every5)
```



```{r quadratic}
quadratic <- function(a,b,c) {
    z<-sqrt(b^2 - (4 * a * c))
    w1= ( b * -1 ) + z
    w2=  ( b * -1 ) - z
    f1=w1/(2*a)
    f2=w2/(2*a)
    print ( f1 )
    print ( f2 )  
}
quadratic(1,0,-9)
```



*I found some interesting functionality to format equations.  This works to display the quadratic root*

$$\sqrt{b^2 - 4ac}$$

        
*But I \*think\* to display the full equation you need to use the \<math\> block*
        
<math xmlns="http://www.w3.org/1998/Math/MathML" display="block">
  <mrow>
    <mi>x</mi>
    <mo>=</mo>
    <mfrac>
      <mrow>
        <mo>&#x2212;</mo>
        <mi>b</mi>
        <mo>&#xB1;</mo>
        <msqrt>
          <mrow>
            <msup>
              <mi>b</mi>
              <mn>2</mn>
            </msup>
            <mo>&#x2212;</mo>
            <mn>4</mn>
            <mi>a</mi>
            <mi>c</mi>
          </mrow>
        </msqrt>
      </mrow>
      <mrow>
        <mn>2</mn>
        <mi>a</mi>
      </mrow>
    </mfrac>
  </mrow>
</math>



