---
title: "CUNY MSDS R Week 2 Assignment"
author: "Tom Buonora"
date: "January 9, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<style type="text/css">
body {
font-size: 16px;
color: black;
}

pre {
  font-size: 16px;
  color: maroon;
}
</style>



<br>
<h1><center>Exercises in Data Frames</center></h1>
<br>

*The data set is a survey of 4700 students who commute to college. 
Attributes include the distance they commute, their high school GPA,
their parents education. <br><br><br>
In this exercise, we look to see if there is a relationship between
parents education and either distance or high school GPA
<br>
The results indicate some correlation between parents education and shorter distances, as well as increased high school GPA.
<br>
The final figures display the outliers of very short and very long distances.

*

<br><br><br>

Issues that I faced :
<ol>
  <li> Couldnt get the gitbhub download to work
  <li> Im trying to integrate some css styles but it sometimes works, sometimes doesnt
</ol>

<br><br><br>


```{r read data}
CollegeDistance.df  <- read.csv("C:\\Users\\arono\\CUNY\\R\\CollegeDistance.csv")


# none of this worked :(

# library (RCurl)
# download <- getURL("http://github.com/TheReallyBigApple/CunyAssignments/blob/main/R/CollegeDistance.csv?accessType=DOWNLOAD")
# CollegeDistance.df <- read.csv (text = download)
# CollegeDistance.df  <- read.csv2("http://github.com/TheReallyBigApple/CunyAssignments/blob/main/R/CollegeDistance.csv")
# tmpFile <- tempfile()
# download.file("http://github.com/TheReallyBigApple/CunyAssignments/blob/main/R/CollegeDistance.csv", destfile = tmpFile, method = "curl")
# url<-"http://github.com/TheReallyBigApple/CunyAssignments/blob/main/R/CollegeDistance.csv"
# url<-"http://raw.github.com/TheReallyBigApple/CunyAssignments/blob/main/R/CollegeDistance.csv"
#  library(RCurl)
# x <- getURL(url)
#  CollegeDistance.df <- read.csv(text = x)

```

<br>


1) Use the summary function to gain an overview of the data set. Then display the mean and median for at least two attributes.


<br>

```{r problem1}
library(knitr)

knitr::kable(summary(CollegeDistance.df, caption='Summary of all Students'))          # score mean = 50.89


score_mean<-round(mean(CollegeDistance.df$score),2) 
score_median<-round(median(CollegeDistance.df$score),2)

distance_mean<-round(mean(CollegeDistance.df$distance),2)
distance_median<-round(median(CollegeDistance.df$distance),2)

matrix1<-matrix(c(score_mean, score_median, distance_mean, distance_median), ncol=2, byrow=TRUE)


rownames(matrix1)=c("Score","Distance")
colnames(matrix1)=c("Mean","Median")

# kable(matrix1,"html")
knitr::kable(matrix1, caption='College Scores and Distances')

```

<br>

2) Create a new data frame with a subset of the columns and rows. Make sure to rename it.
</pre>
<br>


```{r problem2}
# select the gender, ethnicity, score and distance  just the records where the school is in an urban area and both parents went to college
UrbanCollegeEducatedDistance.df <- subset(CollegeDistance.df[c("X","gender", "ethnicity", "score","distance","urban")], CollegeDistance.df$urban == "yes" & 
                                    CollegeDistance.df$fcollege=="yes" & CollegeDistance.df$mcollege=="yes")

```
<br>

3) Create new column names for the new data frame.

<br>
```{r problem3}


# rename the columns to improve their clarity. Note dplyr rename() is neater but, base R is used here

names(UrbanCollegeEducatedDistance.df)[names(UrbanCollegeEducatedDistance.df) == "X"] <- "u_sort"

names(UrbanCollegeEducatedDistance.df)[names(UrbanCollegeEducatedDistance.df) == "score"] <- "u_avg_score_as_HS_Senior"

names(UrbanCollegeEducatedDistance.df)[names(UrbanCollegeEducatedDistance.df) == "gender"] <- "u_gender"

names(UrbanCollegeEducatedDistance.df)[names(UrbanCollegeEducatedDistance.df) == "ethnicity"] <- "u_ethnicity"

names(UrbanCollegeEducatedDistance.df)[names(UrbanCollegeEducatedDistance.df) == "distance"] <- "u_distance"


```
<br> 

4) Use the summary function to create an overview of your new data frame. The print the mean and median for the same two attributes. Please compare.

<br>
```{r problem4}

summary(UrbanCollegeEducatedDistance.df)


score_mean<-round(mean(UrbanCollegeEducatedDistance.df$u_avg_score_as_HS_Senior),2) 
score_median<-round(median(UrbanCollegeEducatedDistance.df$u_avg_score_as_HS_Senior),2)

distance_mean<-round(mean(UrbanCollegeEducatedDistance.df$u_distance),2)
distance_median<-round(median(UrbanCollegeEducatedDistance.df$u_distance),2)

matrix1<-matrix(c(score_mean, score_median, distance_mean, distance_median), ncol=2, byrow=TRUE)

rownames(matrix1)=c("Score","Distance")
colnames(matrix1)=c("Mean","Median")


knitr::kable(matrix1, caption='College Scores and Distances')

```
<br>

5) For at least 3 values in a column please rename so that every value in that column is updated.

<br>
```{r problem5}


# update the sort to 0, so the outliers have u_sort=0

q1 <- quantile(UrbanCollegeEducatedDistance.df$u_distance, .10 )
q3 <- quantile(UrbanCollegeEducatedDistance.df$u_distance, .90 )

sprintf("Isolating students who commuted less than %.2f or greater than %.2f", q1, q3)





UrbanCollegeEducatedDistance.df <- within(UrbanCollegeEducatedDistance.df, {
  f <- u_distance < q1 | u_distance > q3
  u_sort[f] <- 0
}) 





# update other to unknown

UrbanCollegeEducatedDistance.df <- within(UrbanCollegeEducatedDistance.df, {
  f <- u_ethnicity == "other"
  u_ethnicity[f] <- "unknown"
}) 
```
<br>

6) Display enough rows to see examples of all of steps 1-5 above.

<br>
```{r problem6}
# spot check 10 rows in which the sort=0 , these are the outliers
u_outliers<- head(subset(UrbanCollegeEducatedDistance.df,UrbanCollegeEducatedDistance.df$u_sort==0),10)



knitr::kable(u_outliers, caption='College Distance Outliers - Urban Location/Educated Family.')

# spot check 10 rows in which the sort!=0 , these in the first 2 standard deviations ( more or less )
u_norms<-head(subset(UrbanCollegeEducatedDistance.df,UrbanCollegeEducatedDistance.df$u_sort!=0),10)
knitr::kable(u_norms, caption='College Distance Norms - Urban Location/Educated Family.')
```

