---
title: "Data607_Week3"
author: "Tom Buonora"
date: "September 6, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r imports}

library(stringr)    # more consistent than base r
```

<br><br>

1. Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset, provide code that identifies the majors that contain either "DATA" or "STATISTICS"

<br><br>

```{r}

# note: github wraps data in formatting, to get raw data link, click the "Raw" button at the top-left of the data

data_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"
majors<-read.csv(data_file)

str_view(majors$Major, "DATA|STATISTICS", match=TRUE)

```





<br><br>

2. Write code that transforms the data below:
    [1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
    [5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
    [9] "elderberry"   "lime"         "lychee"       "mulberry"    
    [13] "olive"        "salal berry"

<br><br>
Into a format like this:
  c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")
  
<br><br>
  
  
<font  color="purple">The question asks to convert raw data into R code</font>

<br><br>

```{r convert_raw_data_to_r_code}

vector_data<-c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")
dbl_quot<-'\"'
double_quote <- "\"" 
double_quote<-'"'
data_string<-str_c(double_quote,vector_data,double_quote,collapse = ", ")
# data_string<-str_c(vector_data, collapse = ", ") 
code_string<-str_c("c(",data_string,")")
writeLines(code_string)            # writeLines omits the escape characters

```





<br><br>


3. Describe, in words, what these expressions will match:

<br><br>

(.)\1\1

<br><br>

<font  color="purple">Whats inside the parenthesis is a "captured" match. In this case the dot means any character,</font>
<br>
<font  color="purple">The back reference \\1 then takes that match and duplicates a search of the <i>next</i> character only.</font>
<br>
<font  color="purple">i.e. (.)\1\1 will macth any 3 consective characters.</font>
<br><br>

```{r backreference1}

str<-c("BBAAAX", "bBaAAxx", "zzza", "aaaa","...","abaaaaaaax")

str_view(str,"(.)\\1\\1")

```


<br><br>
(.)(.)\\2\\1

<br><br>
<font  color="purple">\\2 will duplicate a search of whats inside the second parenthesis,</font>
<br>
<font  color="purple">thus \\2\\1 effectively will captured match #2 then captured match #1,</font>
<br><br>

```{r}
str<-c("BAAX", "bBaAAxx", "zzza", "abba","...","aaaax")
str_view(str,"(.)(.)\\2\\1")

```

<br><br>
(..)\1
<br><br>
<font  color="purple">Any 2 characters followed the same 2 characters</font>
<br><br>


```{r backreference2}

str<-c("BABAX", "bBaAAxx", "zzza", "abba","...","axaaaxa")
str_view(str,"(..)\1")

```

<br><br>
(.).\\1.\\1
<br><br>
<font  color="purple">Any 2 characters followed the first character followed by any character than the first character again.</font>
<br><br>

```{r backreference3}

str<-c("BABAX", "bBaAAxx", "Z4Z5Z6za", "abba","...","XaXxXaaaxa")
str_view(str,"(.)(.)\\2\\1")


```

<br><br>
"(.)(.)(.).*\\3\\2\\1"
<font  color="purple">Any 3 characters followed by any number of characters we dont care about, followed by </font>
<br>
<font  color="purple">the first 3 characters reversed. Asterix means 0 or more, so the "any number of " </font>
<br>
<font  color="purple">characters in the middle could be zero." </font>
<br><br>

```{r backreference4}
str<-c("BAZAZZA", "OBAZZABOO", "Z4Z5Z6za", "abbba","...","XaXxXaaaXaXxa")
str_view(str,"(.)(.)(.).*\\3\\2\\1")
```


<br><br>

4. Construct regular expressions to match words that:
<br><br>
Start and end with the same character.
<br><br>

```{r start_and_end}

str<-c("BAZAZZA", "OBAZZABOO", "Z4Z5Z6za", "abbba","...","XaXxXaaaXaXxa")
str_view(str,"^(.)\\1$")

```

<br><br>


Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
<br><br>
```{r}
str<-c("BBAAAX", "bBaAAxx", "zzza", "aaaa","...","church")

str_view(str,"(.).*\\1")
```

<br><br>

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

<br><br>



```{r}
str<-c("BBAAAX", "bBaAAxx", "zzza", "aaaNOa","eleven","church")

str_view(str,"(.).*\\1.*\\1")
```
<br><br>
