---
title: "Data607_Week7"
author: "Tom Buonora"
date: "October 8, 2021"
output:
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---







```{r setup, include=FALSE}

knitr::opts_chunk$set(results=TRUE, echo = TRUE, warning = FALSE, message = FALSE)
```



```{r imports and constants, include=FALSE}

library(XML)               # besides xml, this helps us with html by providing the readHTMLTable function
library(tidyverse)         # ggplot2, dplyr, tidyr, readr, tibble, sringr and more
library(knitr)
library(rjson)

CURR_PATH<-str_trim(getwd())

```


<br><br><br>
<font size="7" color="purple">Data Scraping and Parsing : My Favorite Books </font>
<br><br><br>

# Overview


<font size="5" color="blue">  This assignment will take a data set of book data in 3 common formats - html, xml, and json.
<br><br><br>
Then it will parse into a R data frame and display it. </font>
<br><br><br>


<font size="5" color="blue">Each return set is different so standardizing it into a data frame requires some work. </font>
  
  <br><br>

Format        | Function      |  Package   | Returns 
-----------  | -------        | ---------- | ------------- 
HTML         | readHTMLTable  |  XML       | List of Lists    
XML          | xmlParse       |  XML       | XMLInternalDocument    
JSON         | fromJSON       |  rjson     | List of Lists  


<br><br><br>

# HTML

<br><br><br>

<br>
<font size="5" color="blue">  The HTML table format looks like this. 
<br>

  
<pre>

&lt;table&gt; 
 &lt;tr&gt;
    &lt;th&gt;Title&lt;/th&gt;
    &lt;th&gt;Year&lt;/th&gt;
    &lt;th&gt;Type&lt;/th&gt;
    &lt;th&gt;Authors&lt;/th&gt;
  &lt;/tr&gt;
  &lt;tr&gt;
    &lt;td&gt;A Tree Grows in Brooklyn&lt;/td&gt;
    &lt;td&gt;1943&lt;/td&gt;
    &lt;td&gt;Fiction&lt;/td&gt;
    &lt;td&gt;Betty Smith&lt;/td&gt;
  &lt;/tr&gt;
  
  &lt;tr&gt;
    &lt;td&gt;R For Dummies&lt;/td&gt;
    &lt;td&gt;2015&lt;/td&gt;
    &lt;td&gt;Non Fiction&lt;/td&gt;
    &lt;td&gt;Andrie de Vries&lt;/td&gt;&lt;td&gt;Joris Meys&lt;/td&gt;
  &lt;/tr&gt;
  
  &lt;tr&gt;
    &lt;td&gt;The Martian&lt;/td&gt;
    &lt;td&gt;2011&lt;/td&gt;
    &lt;td&gt;Fiction&lt;/td&gt;
    &lt;td&gt;Andy Weir&lt;/td&gt;
  &lt;/tr&gt;
&lt;/table&gt;

 </pre>

<br><br><br>

<br>  <font size="5" color="blue">  Download HTML file.  Parse into dataframe and display. </font>  <br>



```{r parse html table}

html_file<-"https://raw.githubusercontent.com/TheReallyBigApple/CunyAssignments/main/DATA607/MyFavoriteBooks.html"

destfile<-paste0(CURR_PATH,"/books.html")
download.file(html_file, destfile)
tables <- readHTMLTable(destfile)

books_html_df<-as.data.frame(tables[1])

colnames(books_html_df)<-c("Title", " Year", "Type", "Author1", "Author2")


kable(books_html_df, caption="My Favorite Books",row.names = FALSE, booktabs=TRUE)

```

<br><br><br>

# XML


<br><br><br>

<br>
<font size="5" color="blue">  The XML format looks like this. 
<br>



<pre>
&lt;books&gt;
  &lt;book&gt;
    &lt;title&gt;A Tree Grows in Brooklyn&lt;/title&gt; 
    &lt;authors&gt;
      &lt;author&gt;Betty Smith&lt;/author&gt;
    &lt;/authors&gt;
  &lt;year&gt;1943&lt;/year&gt;
   &lt;type&gt;Fiction&lt;/type&gt;
 &lt;/book&gt;
  &lt;book&gt;
    &lt;title&gt;R For Dummies&lt;/title&gt; 
    &lt;authors&gt;
      &lt;author&gt;Joris Meys&lt;/author&gt;
      &lt;author&gt;Andrie de Vries&lt;/author&gt;
    &lt;/authors&gt;
  &lt;year&gt;2015&lt;/year&gt;
   &lt;type&gt;Non Fiction&lt;/type&gt;
 &lt;/book&gt;
  &lt;book&gt;
    &lt;title&gt;The Martian&lt;/title&gt; 
    &lt;authors&gt;
      &lt;author&gt;Andy Weir&lt;/author&gt;
    &lt;/authors&gt;
  &lt;year&gt;2011&lt;/year&gt;
   &lt;type&gt;Fiction&lt;/type&gt;
 &lt;/book&gt;
 &lt;/books&gt; 
 </pre> 

<br><br><br>

<br>  <font size="5" color="blue">  Download XML file.  Parse into dataframe and display. </font>  <br>




```{r xml}


xml_file<-"https://raw.githubusercontent.com/TheReallyBigApple/CunyAssignments/main/DATA607/MyFavoriteBooks.xml"

destfile<-paste0(CURR_PATH,"/books.xml")
download.file(xml_file, destfile)

xml_data<- xmlParse(file = destfile)

xml_df <- xmlToDataFrame(xml_data)

# class(xml_df)
# str(xml_df)
# xml_df

kable(xml_df, caption="My Favorite Books",row.names = FALSE, booktabs=TRUE)

```

<br><br><br>


# JSON


<br><br><br>
<br>
<font size="5" color="blue">  The JSON format looks like this. 
<br>


<pre> 
{"books":[
  { "title":"A Tree Grows in Brooklyn",
   "authors" : [ "Betty Smith" ],
   "year" : "1943",
   "type" : "Fiction"
    },
    { "title":"R For Dummies",
   "authors" : [  "Joris Meys", "Andrie de Vries" ],
   "year" : "2014",
   "type" : "Non Fiction"
    },
    { "title":"The Martian",
   "authors" : [ "Andy Weir" ],
   "year" : "2011",
   "type" : "Fiction"
    }
 
]}
 </pre> 

<br><br><br>

<br>  <font size="5" color="blue">  Download JSON file.  Parse into dataframe and display. </font>  <br>


```{r json}


json_file<-"https://raw.githubusercontent.com/TheReallyBigApple/CunyAssignments/main/DATA607/MyFavoriteBooks.json"

destfile<-paste0(CURR_PATH,"/books.json")
download.file(json_file, destfile)
json_data <- fromJSON(file = "books.json")

# lets look at the structure
# class(json_data)
# str(json_data)


json_df<-as.data.frame(do.call(cbind, json_data$books))
json_df$V2<-as.character(json_df$V2)
json_transposed_df<-t(json_df)

kable(json_transposed_df, caption="My Favorite Books",row.names = FALSE, booktabs=TRUE)



```

<br><br><br>
