---
title: "Data607_Week2"
author: "Tom Buonora"
date: "September 4, 2021"
output: html_document
---


<style type="text/css">
body {
font-size: 18px;
color: black;
}
h1 {
  font-size: 32px;
  color: maroon;
}
h2 {
  font-size: 24px;
  color: black;
}
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```



```{r}

library(RMySQL)
library(dplyr)
library(kableExtra)
library(knitr)
library(tidyr)
```


<br><br>

<h1><center>Movie Ratings : Friends vs. IMDB </center></h1>
<br><br><br>


<br><br>


<h2>This exercise queries a small dataset from MySQL to look at how 5 friends rated 6 movies. </h2>

<h2>The question is how do our friends ratings compare to IMDB. </h2>

<h2>The schema is normalized across 3 tables as follow. Click on link to see the csv data </h2>




<br><br>


  Description       | DB Table  | CSV File
-----------------   | --------  | ---------------
6 movies            | movies    | [607_Movies.csv](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)     
5 people            | people    | [607_People.csv](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)
Movie,Person,Rating | ratings   | [607_Ratings.csv](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)
 


<br><br>

Click [here](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)  to see the database code.


    
The ratings file has 30 entries but some of my friends havent seen the movie yet so they have a blank rating.

We will handle those in the R code.
    

<br><br>

1) Get password from the environment

<br><br>

```{r}

db_password <- Sys.getenv("DATA607_PASSWORD")
```


<br><br>



2) Connect to database


<br><br>

```{r}


conn = dbConnect(MySQL(), dbname='DATA607',  user='root', password=db_password, host='localhost')


```



3) Query the people and movies and ratings tables and show the schema.

```{r}

qry<-"select * from movies;"

rs = dbSendQuery(conn, qry)

  # when n= -1 it means select all rows
movies = fetch(rs, n=-1)

qry<-"select * from people;"

rs = dbSendQuery(conn, qry)

people = fetch(rs, n=-1)

qry<-"select * from ratings;"

rs = dbSendQuery(conn, qry)

ratings = fetch(rs, n=-1)

mhead<-head(movies, n=3)
phead<-head(people, n=3)


```

<br><br>
Movies and People tables :
<br><br>

```{r echo = FALSE}

t1 <- kable(mhead, format = "latex", booktabs = TRUE)
t2 <- kable(phead, format = "latex", booktabs = TRUE)

kable(phead, caption = 'People') %>%
  kable_styling(full_width = FALSE, position = "float_left")

kable(mhead, caption = 'Movies') %>%
  kable_styling(full_width = FALSE, position = "right")
```




```{r echo = FALSE, eval=FALSE}

# save this code
# note:   align = 'clc' means align 3 columns center,left,center

mhead %>%
  kable("html", caption = 'Movies') %>%
    kable_styling(full_width = F, position = "float_left")
phead %>%
  kable("html", caption = 'People') %>%
    kable_styling(full_width = F, position = "center")

```


<br><br>
The Ratings table : 

<br><br>

```{r}

knitr::kable(head(ratings), caption='Ratings', table.attr = "style='width:80%;'")



```


<br><br>

Note we could just run a query like
  select p.name, m.name, r.rating from ratings r, people p, movies m 
    where r.people_id=p.id and r.movie_id=m.id and rating is not null;
    
<br><br>

But we want to practice some R so lets work with dataframes.

<br><br>

1) Remove the records with no rating

<br><br>

```{r}

n_unrated<-nrow(ratings[is.na(ratings$rating),])


sprintf("%d records with no rating will be removed", n_unrated)

ratings<- ratings %>%
  drop_na(rating) 


```

<br><br>

2) Remove the carriage return characters

<br><br>

```{r}

people$name<-trimws(people$name, "right")
movies$lead_actor<-trimws(movies$lead_actor, "right")
```

<br><br>

3) Combine the 3 dataframes to 1
<br><br>

```{r}

ratings2<-merge(people, ratings, by.x ="id", by.y = "people_id")
ratings3<-merge(movies, ratings2, by.x ="id", by.y = "movie_id")

```


<br><br>

4)  Clean up the columns.
<br><br>

```{r}

ratings4<-ratings3 %>%
    select(c("name.y", "name.x", "rating", "imdb_rating","rating"))


colnames(ratings4)<-c("Friend", "Movie", "Rating", "IMDB")
```



<br><br>

5)  Now answer the question.
<br><br>


```{r}

answer<-ratings4 %>%
  group_by(Movie) %>%
  summarize(n_ratings = n(), avg_rating=mean(Rating),IMDB=max(IMDB))

knitr::kable(answer, caption='Friends vs IMDB', digits=1, table.attr = "style='width:80%;'")

```



```{r}

dbDisconnect(conn)


```


The same code was done in Python. Please see [here](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)


