---
title: "Data607_Week5"
author: "Tom Buonora"
date: "September 24, 2021"

output:
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---



<style type="text/css">
body {
font-size: 18px;
color: blue;
}
pre {
  font-size: 12px;
  color: maroon;
}

thead,
tfoot {
    background-color: #3f87a6;
    color: #fff;
}

tbody {
    background-color: #e4f0f5;
}

caption {
    padding: 10px;
    caption-side: bottom;
}

table {
    border-collapse: collapse;
    border: 2px solid rgb(200, 200, 200);
    letter-spacing: 1px;
    font-family: sans-serif;
    font-size: 1.2rem;
}

td,
th {
    border: 1px solid rgb(190, 190, 190);
    padding: 5px 10px;
}

td {
    text-align: center;
}


</style>




```{r setup, include=FALSE}

knitr::opts_chunk$set(results=TRUE, echo = TRUE, warning = FALSE, message = FALSE)
```


<br><br><br>
<font size="7" color="purple">Tidy data using dplyr and tidyr</font>
<br><br><br><br><br><br>

# The R Code

Import readxl and tidyverse
<br>

```{r}

library(readxl)
library(tidyverse)         # ggplot2, dplyr, tidyr, readr, tibble, sringr and more
library(knitr)

```

<br>
Download the excel file. Note the mode is "wb" to preserve the binary elements of the excel file.
<br>

```{r download from internet}

sourcefile<-"https://github.com/acatlin/data/raw/master/israeli_vaccination_data_analysis_start.xlsx"
curr_path<-str_trim(getwd())

destfile<-paste0(curr_path,"/vaccine_data.xlsx")

download.file(sourcefile, destfile, mode = "wb")

```




<br>
Read the excel file into a tibble.
<br>

```{r read_excel}

efile <- read_excel(destfile) 


```


<br>
Get rid of the rows we dont need.
<br>

```{r}

efile <- efile %>% filter(Age == ">50" | Age == "<50") 

```


<br>
Rename the columns.
<br>
```{r rename columns}

efile <- efile %>% rename("pop_nv" = "Population %", "pop_v" = "...3", "sc_nv" = "Severe Cases", "sc_v" = "...5" )

```

<br>
Convert the fields to numeric.
<br>

```{r}
efile<- efile %>% mutate(across(2:6, as.numeric))
```


<br>
Im isolating each individual field as opposed to performing vector operations.
<br>
It provides some clarity and its a good excercise for dplyr functionality.
<br>

```{r define fields}

# overall population
pop_lt_50_nv <- efile %>% filter(Age == "<50") %>% select("pop_nv") %>% first()
pop_gt_50_nv <- efile %>% filter(Age == ">50") %>% select("pop_nv") %>% first()
pop_lt_50_v <- efile %>% filter(Age == "<50") %>% select("pop_v") %>% first()
pop_gt_50_v <- efile %>% filter(Age == ">50") %>% select("pop_v") %>% first()


# severe cases

sc_lt_50_nv<-efile %>% filter(Age == "<50") %>% select("sc_nv") %>% first()
sc_gt_50_nv<-efile %>% filter(Age == ">50") %>% select("sc_nv") %>% first()
sc_lt_50_v<-efile %>% filter(Age == "<50") %>% select("sc_v") %>% first()
sc_gt_50_v<-efile %>% filter(Age == ">50") %>% select("sc_v") %>% first()

```

<br>
Calculate totals.
<br>

```{r}

tot_pop_nv<-pop_lt_50_nv + pop_gt_50_nv
tot_pop_v<-pop_lt_50_v + pop_gt_50_v
tot_pop<-tot_pop_nv+tot_pop_v


tot_sc_nv<-sc_lt_50_nv + sc_gt_50_nv
tot_sc_v<-sc_lt_50_v + sc_gt_50_v
tot_sc<-tot_sc_nv+tot_sc_v
```


<br>
Calculate Attack Rates.
<br>

```{r calculate attack rates}





aru_lt_50<-sc_lt_50_nv / 1000   
aru_gt_50<-sc_gt_50_nv / 1000

arv_lt_50<-sc_lt_50_v / 1000
arv_gt_50<-sc_gt_50_v / 1000

# relative risk is the severe cases of vaccinated over non vaccinated
rr_lt_50 <- arv_lt_50 / aru_lt_50
rr_gt_50 <- arv_gt_50 / aru_gt_50
  
# divide by 2000 since we are averaging the 2 ratios of 1000
aru_all_ages<-(sc_gt_50_nv + sc_lt_50_nv) / 2000
arv_all_ages<-(sc_gt_50_v + sc_lt_50_v) / 2000
rr_all_ages<-(rr_gt_50 + rr_lt_50) / 2

```


<br>
Calculate Vaccine Efficacy.
<br>


```{r calculate efficacy}
ve_lt_50<-(aru_lt_50-arv_lt_50)/aru_lt_50 
ve_gt_50<-(aru_gt_50-arv_gt_50)/aru_gt_50                      


# ve is the difference / aru
ve<-(aru_all_ages-arv_all_ages)/aru_all_ages                   # .674

# same thing
ve<-(1- (arv_all_ages)/(aru_all_ages))



# print(sprintf("The attack rate on the vaccinated for those under 50 is %.3f", arv_lt_50))
# print(sprintf("The attack rate on the vaccinated for those over 50 is %.3f", arv_gt_50))
# print(sprintf("The attack rate for all vaccinated is  %.3f", arv_all_ages))
# 
# print(sprintf("The attack rate on the unvaccinated for those under 50 is %.3f", aru_lt_50))
# print(sprintf("The attack rate on the unvaccinated for those over 50 is %.3f", aru_gt_50))
# print(sprintf("The attack rate for all unvaccinated is %.3f", aru_all_ages))


```

<br>
Add the Efficacies and the Totals to our tibble.
<br>


```{r update tibble}

# efile <- efile %>%  add_column(rr = c(rr_lt_50, rr_gt_50, rr_all_ages),  .after = "sc_v") 
efile <- efile %>%  add_column(rr = 0,  .after = "sc_v") 
efile <-efile %>% rows_insert(tibble("Age"= "Total", pop_nv=tot_pop_nv,  pop_v=tot_pop_v, sc_nv=tot_sc_nv,  sc_v=tot_sc_v, rr=rr_all_ages, Efficacy=ve))
efile <- efile %>% rows_update(tibble(Age= "<50", Efficacy = ve_lt_50))
efile <- efile %>% rows_update(tibble(Age= ">50", Efficacy = ve_gt_50))
efile <- efile %>% rows_update(tibble(Age= "<50", rr = rr_lt_50))
efile <- efile %>% rows_update(tibble(Age= ">50", rr = rr_gt_50))

```

<br>
Round the numerics for presentation purposes.
<br>

```{r}

efile <- efile %>% mutate( Efficacy = round(Efficacy,3 ))

```

<br>
Show everyone our updated tibble.
<br>

```{r show everyone}
kable(efile, caption="",row.names = FALSE, booktabs=TRUE,format.args = list(decimal.mark = '.', big.mark = ",", digits=3))


```


<br>
Save the tibble to a csv file.
<br>
```{r save to csv}
curr_path<-str_trim(getwd())
destfile<-paste0(curr_path,"/vaccine_data_efficacy.csv")
write.csv(efile, destfile)
```

<br><br><br><br><br><br>

# The Questions. The Answers

<br><br><br>


<font size="5" color="maroon">1. What is the total population ?</font>

<font size="4" color="blue"><i>The total population adds up to about 6.9MM. That was Israels total population in 2005
<br>
Its now about 9MM. The percentage values seem to indicate that about 96% of the less than 50 population is accounted for.
<br>
So perhaps this data was created around 2005. If I were to publish analysis on this data, I would need to validate the data.
</i></font>

<br><br><br>

<font size="5" color="maroon">2. Explain the efficacy result.</font>

<font size="4" color="blue"><i>The results indicate that the vaccine adversely effects the older population, i.e. doesnt work.</i></font>

<br><br>
$$Efficacy \ = \ \frac{ARU \ - \ ARV}{ARU} \ * 100%$$
<br>
where 
<br>
$$ARU \ = \ Attack \ Rate \ of \ Unvaccinated \ People \ = \frac{Severe \ Cases (Unvaccinated) }{Population(Unvaccinated) }$$
<br>

$$ARV \ = \ Attack \ Rate \ of \ Vaccinated \ People \ = \frac{Severe \ Cases (Vaccinated) }{Population(Vaccinated) }$$


<br>
<font size="5" color="maroon">(3) Compare the attack rates.</font>
<br><br><br>


<font size="4" color="blue"><i>I dont understand why the severe cases is higher for the over 50 vaccinated group. Is this supposed to represent
an actual distribution to the Isreali population or just a small sample ? I would need to understand the context.
<br>
</i></font>

<br><br><br><br><br><br>

